---
title: "Main Analysis: Ritual markers ~ violent external conflict"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(pacman)
p_load(brms, ape, tidyverse, here, HDInterval, ggdist, tidybayes, ggokabeito, ggthemes)

set.seed(1342)

REFIT <- TRUE

markers <- c(
  "circumcision", "dress", "extra_ritual_group_markers",
  "food_taboos", "hair", "ornaments",
  "permanent_scarring", "tattoos_scarification"
)

model_names <- c("fit_base", "fit_phylo", "fit_phylo_year")

marker_labels <- c(
  "circumcision"               = "Circumcision",
  "dress"                      = "Dress",
  "extra_ritual_group_markers" = "Extra-Ritual In-Group Markers",
  "food_taboos"                = "Food Taboos",
  "hair"                       = "Hair",
  "ornaments"                  = "Ornaments",
  "permanent_scarring"         = "Permanent Scarring",
  "tattoos_scarification"      = "Tattoos or Scarification"
)

model_labels <- c(
  "fit_base"       = "Baseline",
  "fit_phylo"      = "Phylogenetic",
  "fit_phylo_year" = "Phylogenetic + Year + Region"
)

theme_size <- function(base = 13) {
  theme(
    plot.title    = element_text(size = base + 4, face = "bold"),
    plot.subtitle = element_text(size = base + 1),
    axis.title.x  = element_text(size = base + 2),
    axis.title.y  = element_text(size = base + 2),
    axis.text.x   = element_text(size = base),
    axis.text.y   = element_text(size = base),
    legend.title  = element_text(size = base),
    legend.text   = element_text(size = base),
    plot.margin   = margin(10, 14, 10, 14)
  )
}

for (d in c(
  here("data", "model", "external", "fits", "fit_base"),
  here("data", "model", "external", "fits", "fit_phylo"),
  here("data", "model", "external", "fits", "fit_phylo_year"),
  here("data", "model", "external", "results", "fit_base"),
  here("data", "model", "external", "results", "fit_phylo"),
  here("data", "model", "external", "results", "fit_phylo_year"),
  here("figures", "external"),
  here("figures", "external", "region")
)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
```

---

## 1. Fit models

```{r fit_models, message=FALSE, warning=FALSE}
tree <- read.tree(here("data", "jaeger2018", "world.tre"))

fits <- list()

for (marker in markers) {
  data <- read_csv(
    here("data", "model", "external", "input", paste0(marker, ".csv")),
    show_col_types = FALSE
  )

  tips        <- unique(data$tip_name)
  tree_pruned <- keep.tip(tree, tips)
  A           <- vcv.phylo(tree_pruned, corr = TRUE)

  f_base  <- formula(paste(marker, "~ 1 + violent_external"))
  f_phylo <- formula(paste(marker, "~ 1 + violent_external + (1 | gr(tip_name, cov = A))"))
  f_year  <- formula(paste(marker, "~ 1 + violent_external + year_scaled + world_region + (1 | gr(tip_name, cov = A))"))

  prior_base  <- c(prior(normal(0, 1.5), class = Intercept), prior(normal(0, 1), class = b))
  prior_phylo <- c(prior_base, prior(exponential(1), class = sd))

  shared <- list(family = bernoulli(link = logit),
                 chains = 4, iter = 4000, warmup = 2000, cores = 4,
                 backend = "cmdstanr",
                 control = list(adapt_delta = .95, max_treedepth = 15),
                 seed = 1342, silent = 2)

  load_or_fit <- function(model_name, fit_fn) {
    path <- here("data", "model", "external", "fits", model_name, paste0(marker, ".rds"))
    if (!REFIT && file.exists(path)) readRDS(path) else { fit <- fit_fn(); saveRDS(fit, path); fit }
  }

  fits[[marker]] <- list(
    fit_base       = load_or_fit("fit_base", function()
      do.call(brm, c(list(formula = f_base, data = data, prior = prior_base), shared))),
    fit_phylo      = load_or_fit("fit_phylo", function()
      do.call(brm, c(list(formula = f_phylo, data = data, data2 = list(A = A), prior = prior_phylo), shared))),
    fit_phylo_year = load_or_fit("fit_phylo_year", function()
      do.call(brm, c(list(formula = f_year, data = data, data2 = list(A = A), prior = prior_phylo), shared))),
    data           = data,
    A              = A
  )
}

n_labels <- map_chr(markers, function(marker) {
  paste0(marker_labels[marker], " (n=", nrow(fits[[marker]]$data), ")")
}) |> setNames(markers)
```

---

## 2. Extract and save results

Saves per-model subfolders: coefficients (fixed effects, log-odds scale), hypothesis tests
(P(violent_external > 0)), and MCMC diagnostics (Rhat, ESS).

```{r extract_results}
for (mname in model_names) {
  fe <- map_dfr(markers, function(marker) {
    draws <- fixef(fits[[marker]][[mname]], summary = FALSE)
    map_dfr(colnames(draws), function(param) {
      d <- draws[, param]
      tibble(marker = marker, parameter = param,
             median = median(d), lo95 = hdi(d)[1], hi95 = hdi(d)[2])
    })
  }) |> mutate(across(where(is.numeric), \(x) round(x, 2)))
  write_csv(fe, here("data", "model", "external", "results", mname, "coefficients.csv"))

  ht <- map_dfr(markers, function(marker) {
    h <- hypothesis(fits[[marker]][[mname]], "violent_external > 0")$hypothesis
    tibble(marker = marker,
           estimate   = h$Estimate,
           lo95       = h$CI.Lower,
           hi95       = h$CI.Upper,
           post_prob  = h$Post.Prob,
           evid_ratio = h$Evid.Ratio)
  }) |> mutate(across(where(is.numeric), \(x) round(x, 2)))
  write_csv(ht, here("data", "model", "external", "results", mname, "hypothesis_tests.csv"))

  diag <- map_dfr(markers, function(marker) {
    s <- summary(fits[[marker]][[mname]])$fixed
    tibble(marker = marker, parameter = rownames(s),
           rhat = s[, "Rhat"], ess_bulk = s[, "Bulk_ESS"], ess_tail = s[, "Tail_ESS"])
  }) |> mutate(across(where(is.numeric), \(x) round(x, 2)))
  write_csv(diag, here("data", "model", "external", "results", mname, "diagnostics.csv"))
}

# Flag any convergence issues across all models
map_dfr(model_names, function(mname) {
  read_csv(here("data", "model", "external", "results", mname, "diagnostics.csv"),
           show_col_types = FALSE) |> mutate(model = mname)
}) |> filter(rhat > 1.01 | ess_bulk < 400 | ess_tail < 400)
```

---

## 3. Forest plot: violent_external (log-odds, all models)

Point estimates (posterior median) and 95% HDI for `b_violent_external` across all three models.
Shows robustness of the effect to phylogenetic and covariate adjustment.

```{r forest_logodds}
coef_all <- map_dfr(model_names, function(mname) {
  read_csv(here("data", "model", "external", "results", mname, "coefficients.csv"),
           show_col_types = FALSE) |> mutate(model = mname)
}) |>
  filter(parameter == "violent_external") |>
  mutate(
    marker_label = factor(n_labels[marker], levels = rev(n_labels[markers])),
    model_label  = factor(model_labels[model], levels = model_labels)
  )

p <- ggplot(coef_all, aes(x = median, xmin = lo95, xmax = hi95,
                           y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Coefficient (log-odds)", y = NULL, color = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "Posterior median and 95% HDI") +
  theme_bw() +
  theme_size() +
  theme(legend.position = "bottom")

ggsave(here("figures", "external", "forest_logodds.png"), p, width = 10, height = 7)
p
```

---

## 4. Posterior distributions: violent_external (fit_phylo_year)

Marker ordering by median log-odds effect (ascending) is fixed and shared across all three plots below.

```{r marker_order}
draws_logodds <- map_dfr(markers, function(marker) {
  tibble(marker = marker,
         draw   = as_draws_df(fits[[marker]]$fit_phylo_year)$b_violent_external)
})

marker_order <- draws_logodds |>
  group_by(marker) |>
  summarise(med = median(draw), .groups = "drop") |>
  arrange(med) |>
  pull(marker)
```

### Plot 1: log-odds

The raw coefficient from the model. The natural scale for hypothesis testing.
A positive value means conflict is associated with higher log-odds of the marker.

```{r betas_logodds}
p <- draws_logodds |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Coefficient (log-odds)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "fit_phylo_year — posterior distribution (log-odds)") +
  theme_bw() +
  theme_size()

ggsave(here("figures", "external", "betas_logodds.png"), p, width = 9, height = 7)
p
```

### Plot 2: AME (probability scale)

Average marginal effect: for each posterior draw, compute
P(Y=1 | conflict=1, covariates) − P(Y=1 | conflict=0, covariates),
averaged equally over all observed world regions at mean year (re_formula = NA).
This is the correct probability-scale representation of the effect — not plogis(beta),
which would ignore the actual baseline probability.

```{r betas_ame, message=FALSE, warning=FALSE}
grand_mean_draws <- map_dfr(markers, function(marker) {
  fit  <- fits[[marker]]$fit_phylo_year
  data <- fits[[marker]]$data

  nd <- expand_grid(
    violent_external = c(0, 1),
    year_scaled      = 0,
    world_region     = unique(data$world_region)
  )

  epred_draws(fit, newdata = nd, re_formula = NA) |>
    ungroup() |>
    mutate(marker           = marker,
           violent_external = factor(violent_external, levels = c(0, 1),
                                     labels = c("No conflict", "Conflict"))) |>
    group_by(marker, .draw, violent_external) |>
    summarise(.epred = mean(.epred), .groups = "drop")
})

ame_draws <- grand_mean_draws |>
  group_by(marker, .draw) |>
  summarise(ame = .epred[violent_external == "Conflict"] - .epred[violent_external == "No conflict"],
            .groups = "drop")

p <- ame_draws |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = ame, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Average marginal effect (probability)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "fit_phylo_year — AME at mean year, averaged over regions") +
  theme_bw() +
  theme_size()

ggsave(here("figures", "external", "betas_ame.png"), p, width = 9, height = 7)
p
```

### Plot 3: Grand mean predictions (both conditions)

Shows the full posterior predictive distribution for both conditions — not just the difference
but the absolute predicted probability for each. Reveals baseline prevalence alongside the shift.
A large log-odds coefficient can still mean a small probability shift if baseline prevalence is low.

```{r grand_mean_predictions}
p <- grand_mean_draws |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = .epred, y = marker_label, fill = violent_external)) +
  stat_halfeye(alpha = 0.9, normalize = "groups", adjust = 0.8) +
  scale_fill_okabe_ito(name = "Violent external conflict") +
  labs(x = "Predicted probability", y = NULL,
       title = "Predicted probability of marker by conflict condition",
       subtitle = "fit_phylo_year — population-level, mean year, averaged over regions") +
  theme_clean() +
  theme_size() +
  theme(legend.position = "bottom")

ggsave(here("figures", "external", "grand_mean_predictions.png"), p, width = 9.5, height = 6.5)
p
```

---

## 5. Year effect (fit_phylo_year)

Posterior distribution of `b_year_scaled` — the log-odds change per one SD increase in year.
Year is z-scored within each marker's dataset, so one unit = one SD of the year variable.

```{r year_effect}
draws_year <- map_dfr(markers, function(marker) {
  tibble(marker = marker,
         draw   = as_draws_df(fits[[marker]]$fit_phylo_year)$b_year_scaled)
})

year_order <- draws_year |>
  group_by(marker) |>
  summarise(med = median(draw), .groups = "drop") |>
  arrange(med) |>
  pull(marker)

p <- draws_year |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[year_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#7BAFD4", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Coefficient (log-odds per SD of year)", y = NULL,
       title = "Effect of year on ritual markers",
       subtitle = "fit_phylo_year — posterior distribution") +
  theme_bw() +
  theme_size()

ggsave(here("figures", "external", "year_effect.png"), p, width = 9, height = 7)
p
```

---

## 6. Region baseline distributions (fit_phylo_year)

For each marker: the absolute baseline log-odds (and probability) per world region,
at `violent_external = 0` and `year_scaled = 0`. Each region's distribution is computed
draw-by-draw as `b_Intercept + b_world_region*`, with Africa (the reference level) using
`b_Intercept` alone. This correctly propagates the joint posterior uncertainty of both parameters.

On the probability scale, `plogis(b_Intercept + b_world_region*)` gives the predicted
probability of the marker for a society in that region with no conflict at the mean year.

```{r region_betas, message=FALSE, warning=FALSE}
scales <- list(
  logodds     = list(
    transform = identity,
    xlab      = "Log-odds (absolute baseline, violent_external = 0, year = mean)",
    subtitle  = "World region baseline log-odds — fit_phylo_year"
  ),
  probability = list(
    transform = plogis,
    xlab      = "Predicted probability (violent_external = 0, year = mean)",
    subtitle  = "World region baseline probability — fit_phylo_year"
  )
)

for (marker in markers) {
  fit <- fits[[marker]]$fit_phylo_year

  # Join on 7-char prefix to handle Oceania label mismatch between old fits and current data
  # ("OceaniaMAustralia" and "OceaniaAustralia" both share prefix "Oceania"; all other
  # regions are uniquely distinguishable at 7 characters)
  n_region <- as_tibble(fit$data) |>
    mutate(world_region = as.character(world_region)) |>
    count(world_region) |>
    mutate(param_prefix = substr(gsub("[^[:alnum:]]", "", world_region), 1, 7))

  africa_n <- n_region |> filter(world_region == "Africa") |> pull(n)

  draws_df    <- as_draws_df(fit)
  region_cols <- grep("^b_world_region", names(draws_df), value = TRUE)

  draws_long <- draws_df |>
    select(all_of(region_cols), .draw, b_Intercept) |>
    pivot_longer(-c(.draw, b_Intercept), names_to = "param", values_to = "beta") |>
    mutate(
      param_key    = str_remove(param, "^b_world_region"),
      param_prefix = substr(param_key, 1, 7),
      value        = b_Intercept + beta
    ) |>
    left_join(n_region |> select(world_region, n, param_prefix), by = "param_prefix")

  africa_draws <- draws_df |>
    transmute(
      .draw        = .draw,
      param        = "b_world_regionAfrica",
      value        = b_Intercept,
      param_key    = "Africa",
      param_prefix = "Africa",
      world_region = "Africa",
      n            = africa_n
    )

  draws_all <- bind_rows(draws_long, africa_draws) |>
    mutate(region_label = paste0(world_region, " (n=", n, ")"))

  region_order <- draws_all |>
    group_by(region_label) |>
    summarise(med = median(value), .groups = "drop") |>
    arrange(med) |>
    pull(region_label)

  draws_all <- draws_all |>
    mutate(
      region_label = factor(region_label, levels = region_order),
      is_africa    = world_region == "Africa"
    )

  for (scale_name in names(scales)) {
    sc <- scales[[scale_name]]

    d <- draws_all |> mutate(value = sc$transform(value))

    xlims <- quantile(d$value, c(0.005, 0.995))

    p <- ggplot(d, aes(x = value, y = region_label, fill = is_africa)) +
      stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi, alpha = 0.85) +
      scale_fill_manual(values = c("FALSE" = "#5B8DB8", "TRUE" = "grey65"), guide = "none") +
      labs(x = sc$xlab, y = NULL,
           title    = n_labels[marker],
           subtitle = sc$subtitle) +
      coord_cartesian(xlim = xlims) +
      theme_bw() +
      theme_size()

    ggsave(
      here("figures", "external", "region", paste0("region_betas_", scale_name, "_", marker, ".png")),
      p, width = 8, height = 6
    )
  }
}
```
