---
title: "Supplementary Analysis: Ritual markers ~ violent external conflict (eHRAF excluded)"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(pacman)
p_load(brms, ape, tidyverse, here, HDInterval, marginaleffects, loo, ggdist, tidybayes, ggokabeito, ggthemes)

set.seed(1342)

REFIT <- TRUE  # set TRUE to refit all models from scratch

markers <- c(
  "circumcision", "dress", "extra_ritual_group_markers",
  "food_taboos", "hair", "ornaments",
  "permanent_scarring", "tattoos_scarification"
)

model_names <- c("fit_base", "fit_phylo", "fit_phylo_repeat", "fit_phylo_year")

marker_labels <- c(
  "circumcision"               = "Circumcision",
  "dress"                      = "Dress",
  "extra_ritual_group_markers" = "Extra-Ritual In-Group Markers",
  "food_taboos"                = "Food Taboos",
  "hair"                       = "Hair",
  "ornaments"                  = "Ornaments",
  "permanent_scarring"         = "Permanent Scarring",
  "tattoos_scarification"      = "Tattoos or Scarification"
)

model_labels <- c(
  "fit_base"         = "Baseline",
  "fit_phylo"        = "Phylogenetic",
  "fit_phylo_repeat" = "Phylogenetic + Repeated",
  "fit_phylo_year"   = "Phylogenetic + Year + Region"
)

# Shared theme for grand mean / empirical plots (matches 4_plot_main.R style)
theme_size <- function(base = 13) {
  theme(
    plot.title   = element_text(size = base + 4, face = "bold"),
    plot.subtitle = element_text(size = base + 1),
    axis.title.x = element_text(size = base + 2),
    axis.title.y = element_text(size = base + 2),
    axis.text.x  = element_text(size = base),
    axis.text.y  = element_text(size = base),
    legend.title = element_text(size = base),
    legend.text  = element_text(size = base),
    plot.margin  = margin(10, 14, 10, 14)
  )
}

# Output directories
for (d in c(
  here("data", "model", "external_noeHRAF", "fits", "fit_base"),
  here("data", "model", "external_noeHRAF", "fits", "fit_phylo"),
  here("data", "model", "external_noeHRAF", "fits", "fit_phylo_repeat"),
  here("data", "model", "external_noeHRAF", "fits", "fit_phylo_year"),
  here("data", "model", "external_noeHRAF", "fits"),
  here("data", "model", "external_noeHRAF", "results"),
  here("figures", "external_noeHRAF")
)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

for (m in markers)
  dir.create(here("figures", "external_noeHRAF", "diagnostics", m), recursive = TRUE, showWarnings = FALSE)
```

---

## 1. Fit models

Set `REFIT <- TRUE` to refit from scratch; otherwise loads cached `.rds` files.

NB: consider whether we should fit model 3 at all (f_rep)

```{r fit_models, message=FALSE, warning=FALSE}
tree <- read.tree(here("data", "jaeger2018", "world.tre"))

fits <- list()

for (marker in markers) {

  data <- read_csv(
    here("data", "model", "external_noeHRAF", "input", paste0(marker, ".csv")),
    show_col_types = FALSE
  ) |> mutate(tip_name_rep = tip_name)

  tips        <- unique(data$tip_name)
  tree_pruned <- keep.tip(tree, tips)
  A           <- vcv.phylo(tree_pruned, corr = TRUE)

  f_base  <- formula(paste(marker, "~ 1 + violent_external"))
  f_phylo <- formula(paste(marker, "~ 1 + violent_external + (1 | gr(tip_name, cov = A))"))
  f_rep   <- formula(paste(marker, "~ 1 + violent_external + (1 | gr(tip_name, cov = A)) + (1 | tip_name_rep)"))
  f_year  <- formula(paste(marker, "~ 1 + violent_external + year_scaled + world_region + (1 | gr(tip_name, cov = A))"))

  prior_base  <- c(prior(normal(0, 1.5), class = Intercept), prior(normal(0, 1), class = b))
  prior_phylo <- c(prior_base, prior(exponential(1), class = sd))

  shared <- list(family = bernoulli(link = logit),
                 chains = 4, iter = 4000, warmup = 2000, cores = 4,
                 backend = "cmdstanr",
                 control = list(adapt_delta = .95, max_treedepth = 15),
                 seed = 1342, silent = 2)

  load_or_fit <- function(model_name, fit_fn) {
    path <- here("data", "model", "external_noeHRAF", "fits", model_name, paste0(marker, ".rds"))
    if (!REFIT && file.exists(path)) {
      readRDS(path)
    } else {
      fit <- fit_fn()
      saveRDS(fit, path)
      fit
    }
  }

  fit_base <- load_or_fit("fit_base", function()
    do.call(brm, c(list(formula = f_base, data = data, prior = prior_base), shared)))

  fit_phylo <- load_or_fit("fit_phylo", function()
    do.call(brm, c(list(formula = f_phylo, data = data, data2 = list(A = A), prior = prior_phylo), shared)))

  fit_phylo_repeat <- load_or_fit("fit_phylo_repeat", function()
    do.call(brm, c(list(formula = f_rep, data = data, data2 = list(A = A), prior = prior_phylo), shared)))

  fit_phylo_year <- load_or_fit("fit_phylo_year", function()
    do.call(brm, c(list(formula = f_year, data = data, data2 = list(A = A), prior = prior_phylo), shared)))

  fits[[marker]] <- list(
    fit_base         = fit_base,
    fit_phylo        = fit_phylo,
    fit_phylo_repeat = fit_phylo_repeat,
    fit_phylo_year   = fit_phylo_year,
    data             = data,
    A                = A
  )
}

# Marker labels with sample sizes
n_labels <- map_chr(markers, function(marker) {
  paste0(marker_labels[marker], " (n=", nrow(fits[[marker]]$data), ")")
}) |> setNames(markers)
```

---

## 2. Extract results

### Fixed effects (log-odds scale)

```{r fixed_effects}
fixed_effects <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    fe <- fixef(fits[[marker]][[mname]], summary = FALSE)
    map_dfr(colnames(fe), function(param) {
      draws <- fe[, param]
      tibble(marker = marker, model = mname, parameter = param,
             median = median(draws), lo95 = hdi(draws)[1], hi95 = hdi(draws)[2])
    })
  })
})

fixed_effects <- fixed_effects |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(fixed_effects, here("data", "model", "external_noeHRAF", "results", "coefficients.csv"))
fixed_effects |> filter(parameter == "violent_external")
```

### Hypothesis tests: P(violent_external > 0)

```{r hypothesis_tests}
hypothesis_tests <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    h <- hypothesis(fits[[marker]][[mname]], "violent_external > 0")$hypothesis
    tibble(marker = marker, model = mname,
           estimate   = h$Estimate,
           lo95       = h$CI.Lower,
           hi95       = h$CI.Upper,
           post_prob  = h$Post.Prob,
           evid_ratio = h$Evid.Ratio)
  })
})

hypothesis_tests <- hypothesis_tests |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(hypothesis_tests, here("data", "model", "external_noeHRAF", "results", "hypothesis_tests.csv"))
hypothesis_tests
```

### Natural scale: OR and AME

OR = exp(b_violent_external): multiplicative change in odds.

The odds of a marker being present (e.g. tattoos) are X times higher in societies with violent external conflict than in societies without it, holding everything else constant. E.g., OR=2 means double, but applies everywhere one scale (e.g., 1%-->2% or 10%-->20%).

AME = average change in P(outcome = 1) for violent_external 0 → 1, averaged over the data.

Result is in probability units e.g., 0.12 means that, on average across the sample, having violent external conflict is associated with 12 percentage-point higher probability of marker being present. AME accounts for the nonlinearity of the logistic function — the same log-odds coefficient translates to a larger probability shift near P=0.5 and a smaller shift near P=0 or P=1. The AME averages over where each observation actually sits on that curve, so it's more directly interpretable as a real-world effect size. (read more up on this.)

```{r natural_scale}
natural_scale <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    fit  <- fits[[marker]][[mname]]
    data <- fits[[marker]]$data

    # Odds ratio from posterior draws
    draws    <- as_draws_df(fit)$b_violent_external
    or_draws <- exp(draws)

    # Average marginal effect (probability scale)
    ame <- avg_slopes(fit, variables = "violent_external", newdata = data)

    tibble(
      marker     = marker,
      model      = mname,
      or_median  = median(or_draws),
      or_lo95    = hdi(or_draws)[1],
      or_hi95    = hdi(or_draws)[2],
      ame_median = ame$estimate,
      ame_lo95   = ame$conf.low,
      ame_hi95   = ame$conf.high
    )
  })
})

natural_scale <- natural_scale |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(natural_scale, here("data", "model", "external_noeHRAF", "results", "natural_scale.csv"))
natural_scale
```

### Phylogenetic signal (λ)

```{r phylo_signal}
phylo_signal <- map_dfr(markers, function(marker) {
  # fit_phylo
  h1 <- hypothesis(
    fits[[marker]]$fit_phylo,
    "sd_tip_name__Intercept^2 / (sd_tip_name__Intercept^2 + 3.289868) = 0",
    class = NULL
  )$hypothesis

  # fit_phylo_repeat
  h2 <- hypothesis(
    fits[[marker]]$fit_phylo_repeat,
    paste("sd_tip_name__Intercept^2 /",
          "(sd_tip_name__Intercept^2 + sd_tip_name_rep__Intercept^2 + 3.289868) = 0"),
    class = NULL
  )$hypothesis

  bind_rows(
    tibble(marker = marker, model = "fit_phylo",        lambda = h1$Estimate, lo95 = h1$CI.Lower, hi95 = h1$CI.Upper),
    tibble(marker = marker, model = "fit_phylo_repeat", lambda = h2$Estimate, lo95 = h2$CI.Lower, hi95 = h2$CI.Upper)
  )
})

phylo_signal <- phylo_signal |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(phylo_signal, here("data", "model", "external_noeHRAF", "results", "phylo_signal.csv"))
phylo_signal
```

### Diagnostics (Rhat, ESS)

```{r diagnostics}
diagnostics <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    s <- summary(fits[[marker]][[mname]])$fixed
    tibble(marker = marker, model = mname, parameter = rownames(s),
           rhat = s[, "Rhat"], ess_bulk = s[, "Bulk_ESS"], ess_tail = s[, "Tail_ESS"])
  })
})

diagnostics <- diagnostics |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(diagnostics, here("data", "model", "external_noeHRAF", "results", "diagnostics.csv"))

# Flag any potential issues
diagnostics |> filter(rhat > 1.01 | ess_bulk < 400 | ess_tail < 400)
```

---

## 3. LOO model comparison

```{r loo_comparison, message=FALSE, warning=FALSE}
loo_results <- map_dfr(markers, function(marker) {
  loos <- list(
    fit_base         = loo(fits[[marker]]$fit_base),
    fit_phylo        = loo(fits[[marker]]$fit_phylo),
    fit_phylo_repeat = loo(fits[[marker]]$fit_phylo_repeat),
    fit_phylo_year   = loo(fits[[marker]]$fit_phylo_year)
  )
  comparison <- loo_compare(x = loos)
  as.data.frame(comparison) |>
    rownames_to_column("model") |>
    mutate(marker = marker)
})

loo_results <- loo_results |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(loo_results, here("data", "model", "external_noeHRAF", "results", "loo_comparison.csv"))
loo_results
```

---

## 4. Summary forest plots (all models)

Point + range plots showing all four models per marker.

```{r forest_logodds}
coef_ve <- fixed_effects |>
  filter(parameter == "violent_external") |>
  mutate(
    marker_label = factor(n_labels[marker], levels = n_labels[markers]),
    model_label  = factor(model_labels[model], levels = model_labels)
  )

p <- ggplot(coef_ve, aes(x = median, xmin = lo95, xmax = hi95,
                          y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Coefficient (log-odds)", y = NULL, color = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "All models — posterior median and 95% HDI") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "external_noeHRAF", "forest_logodds.png"), p, width = 10, height = 7)
p
```

```{r forest_or}
ns_plot <- natural_scale |>
  mutate(
    marker_label = factor(n_labels[marker], levels = n_labels[markers]),
    model_label  = factor(model_labels[model], levels = model_labels)
  )

p <- ggplot(ns_plot, aes(x = or_median, xmin = or_lo95, xmax = or_hi95,
                          y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  labs(x = "Odds Ratio (log scale)", y = NULL, color = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "All models — odds ratios") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "external_noeHRAF", "forest_or.png"), p, width = 10, height = 7)
p
```

```{r forest_ame}
p <- ggplot(ns_plot, aes(x = ame_median, xmin = ame_lo95, xmax = ame_hi95,
                          y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Average Marginal Effect (probability scale)", y = NULL, color = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "All models — AME") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "external_noeHRAF", "forest_ame.png"), p, width = 10, height = 7)
p
```

---

## 5. Main model distribution plots (fit_phylo_year)

Posterior distributions for `fit_phylo_year` only. Markers ordered by median effect size.

```{r m4_draws}
# Log-odds posterior draws from fit_phylo_year
draws_logodds <- map_dfr(markers, function(marker) {
  draws <- as_draws_df(fits[[marker]]$fit_phylo_year)$b_violent_external
  tibble(marker = marker, draw = draws)
})

# OR draws
draws_or <- draws_logodds |> mutate(draw = exp(draw))

# AME draws (full posterior distribution)
draws_ame <- map_dfr(markers, function(marker) {
  fit  <- fits[[marker]]$fit_phylo_year
  data <- fits[[marker]]$data
  pd   <- posterior_draws(avg_slopes(fit, variables = "violent_external", newdata = data))
  tibble(marker = marker, draw = pd$draw)
})

# Marker order by median log-odds effect (ascending, so largest is on top)
marker_order <- draws_logodds |>
  group_by(marker) |>
  summarise(med = median(draw), .groups = "drop") |>
  arrange(med) |>
  pull(marker)
```

```{r m4_dist_logodds}
p <- draws_logodds |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Coefficient (log-odds)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "Phylogenetic + Year + Region — posterior distribution") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 11))

ggsave(here("figures", "external_noeHRAF", "m4_dist_logodds.png"), p, width = 9, height = 7)
p
```

```{r m4_dist_or}
p <- draws_or |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  labs(x = "Odds Ratio (log scale)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "Phylogenetic + Year + Region — posterior distribution") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 11))

ggsave(here("figures", "external_noeHRAF", "m4_dist_or.png"), p, width = 9, height = 7)
p
```

```{r m4_dist_ame}
# Use same marker order as log-odds
ame_order <- draws_ame |>
  group_by(marker) |>
  summarise(med = median(draw), .groups = "drop") |>
  arrange(med) |>
  pull(marker)

p <- draws_ame |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[ame_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Average Marginal Effect (probability scale)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "Phylogenetic + Year + Region — posterior distribution") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 11))

ggsave(here("figures", "external_noeHRAF", "m4_dist_ame.png"), p, width = 9, height = 7)
p
```

## 6. Main model: grand mean and empirical predictions

Two complementary probability-scale views of `fit_phylo_year`. Both use `re_formula = NA`
(population-level, stripping the phylogenetic RE) so predictions are not tied to specific language groups.

- **Grand mean**: predictions at `year_scaled = 0`, averaged across all observed world regions.
  Cleanest view of the pure fixed effect.
- **Empirical**: averaged over actual observed `(world_region, year_scaled)` rows in each marker's data.
  More honest about the covariate distribution in the sample.

If grand mean AME ≈ empirical AME, results are robust to the choice. For the main text, the grand mean
AME is the most direct plot; the two-condition version shows absolute prevalence alongside the shift.

```{r grand_mean_draws, message=FALSE, warning=FALSE}
grand_mean_draws <- map_dfr(markers, function(marker) {
  fit  <- fits[[marker]]$fit_phylo_year
  data <- fits[[marker]]$data

  nd <- expand_grid(
    violent_external = c(0, 1),
    year_scaled      = 0,
    world_region     = unique(data$world_region)
  )

  epred_draws(fit, newdata = nd, re_formula = NA) |>
    ungroup() |>
    mutate(
      marker           = marker,
      violent_external = factor(violent_external, levels = c(0, 1), labels = c("No", "Yes"))
    ) |>
    group_by(marker, .draw, violent_external) |>
    summarise(.epred = mean(.epred), .groups = "drop")
})

grand_ame_draws <- grand_mean_draws |>
  group_by(marker, .draw) |>
  summarise(ame = .epred[violent_external == "Yes"] - .epred[violent_external == "No"],
            .groups = "drop")
```

```{r empirical_draws, message=FALSE, warning=FALSE}
empirical_draws <- map_dfr(markers, function(marker) {
  fit  <- fits[[marker]]$fit_phylo_year
  data <- fits[[marker]]$data

  set.seed(1342)
  nd_cov <- data |>
    select(world_region, year_scaled) |>
    slice_sample(n = min(300, nrow(data)))
  nd <- expand_grid(violent_external = c(0, 1), nd_cov)

  epred_draws(fit, newdata = nd, re_formula = NA) |>
    ungroup() |>
    mutate(
      marker           = marker,
      violent_external = factor(violent_external, levels = c(0, 1), labels = c("No", "Yes"))
    ) |>
    group_by(marker, .draw, violent_external) |>
    summarise(.epred = mean(.epred), .groups = "drop")
})

empirical_ame_draws <- empirical_draws |>
  group_by(marker, .draw) |>
  summarise(ame = .epred[violent_external == "Yes"] - .epred[violent_external == "No"],
            .groups = "drop")
```

```{r plot_grand_mean_predictions}
# Marker order by grand mean AME
gm_order <- grand_ame_draws |>
  group_by(marker) |>
  summarise(med = median(ame), .groups = "drop") |>
  arrange(med) |>
  pull(marker)

p <- grand_mean_draws |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[gm_order])) |>
  ggplot(aes(x = .epred, y = marker_label, fill = violent_external)) +
  stat_halfeye(alpha = 0.9, normalize = "groups", adjust = 0.8) +
  scale_fill_okabe_ito(name = "Violent external conflict") +
  labs(x = "Expected probability (E[Y | X])", y = NULL,
       title = "Posterior mean probability by marker",
       subtitle = "Population-level (no phylo RE), year scaled = 0, averaged over regions") +
  theme_clean() +
  theme_size(base = 13) +
  theme(legend.position = "bottom")

ggsave(here("figures", "external_noeHRAF", "m4_grand_mean_predictions.png"), p, width = 9.5, height = 6.5)
p
```

```{r plot_grand_ame}
p <- grand_ame_draws |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[gm_order])) |>
  ggplot(aes(x = ame, y = marker_label)) +
  stat_halfeye(alpha = 0.95, adjust = 0.8, fill = palette_okabe_ito(order = 7),
               .width = c(0.66, 0.95), point_interval = median_hdi) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "AME in expected probability (E[Y|Yes] - E[Y|No])", y = NULL,
       title = "Average marginal effect (Yes - No) by marker",
       subtitle = "Population-level (no phylo RE), year scaled = 0, averaged over regions") +
  theme_clean() +
  theme_size(base = 13)

ggsave(here("figures", "external_noeHRAF", "m4_grand_ame.png"), p, width = 9.5, height = 6.5)
p
```

```{r plot_empirical_ame}
emp_order <- empirical_ame_draws |>
  group_by(marker) |>
  summarise(med = median(ame), .groups = "drop") |>
  arrange(med) |>
  pull(marker)

p <- empirical_ame_draws |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[emp_order])) |>
  ggplot(aes(x = ame, y = marker_label)) +
  stat_halfeye(alpha = 0.95, adjust = 0.8, fill = palette_okabe_ito(order = 7),
               .width = c(0.66, 0.95), point_interval = median_hdi) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "AME in expected probability (E[Y|Yes] - E[Y|No])", y = NULL,
       title = "Average marginal effect (Yes - No) by marker",
       subtitle = "Empirical: averaged over observed region × year distribution") +
  theme_clean() +
  theme_size(base = 13)

ggsave(here("figures", "external_noeHRAF", "m4_empirical_ame.png"), p, width = 9.5, height = 6.5)
p
```

---

## 7. Phylogenetic signal plot

```{r phylo_signal_plot}
p <- phylo_signal |>
  mutate(
    marker_label = factor(n_labels[marker], levels = n_labels[markers]),
    model_label  = factor(model_labels[model], levels = model_labels)
  ) |>
  ggplot(aes(x = lambda, xmin = lo95, xmax = hi95, y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Phylogenetic signal (λ)", y = NULL, color = NULL,
       title = "Phylogenetic signal across markers",
       subtitle = "Posterior median and 95% HDI") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "external_noeHRAF", "phylo_signal.png"), p, width = 10, height = 7)
p
```

---

## 8. Exploratory: AME by region

For each marker: posterior AME of `violent_external` broken down by `world_region`,
at `year_scaled = 0`, population-level (`re_formula = NA`). Useful for understanding
whether the effect is driven by particular regions. Saved to `figures/diagnostics/{marker}/`.

```{r region_ame, message=FALSE, warning=FALSE}
for (marker in markers) {
  fit  <- fits[[marker]]$fit_phylo_year
  data <- fits[[marker]]$data

  nd <- expand_grid(
    violent_external = c(0, 1),
    year_scaled      = 0,
    world_region     = unique(data$world_region)
  )

  region_ame <- epred_draws(fit, newdata = nd, re_formula = NA) |>
    ungroup() |>
    mutate(violent_external = factor(violent_external, levels = c(0, 1), labels = c("No", "Yes"))) |>
    group_by(.draw, world_region) |>
    summarise(ame = .epred[violent_external == "Yes"] - .epred[violent_external == "No"],
              .groups = "drop")

  p <- region_ame |>
    ggplot(aes(x = ame, y = reorder(world_region, ame, median))) +
    stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
                 fill = "#C8839E", color = "black", alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "AME (probability scale)", y = NULL,
         title = paste("AME by region:", n_labels[marker]),
         subtitle = "Phylogenetic + Year + Region — year scaled = 0, population-level") +
    theme_bw()

  ggsave(here("figures", "external_noeHRAF", "diagnostics", marker, "region_ame.png"), p, width = 8, height = 6)
}
```

---

## Notes: AME approaches compared

There are several different AME-related quantities computed in this script. They answer
slightly different questions and are worth comparing carefully before deciding what goes
in the paper.

### 1. `avg_slopes` AME — section 5 (`m4_dist_ame`)

**How**: `marginaleffects::avg_slopes(fit, variables = "violent_external", newdata = data)`,
then `posterior_draws()` to get the full distribution.

**What it averages over**: the actual observed rows in the data (all `world_region` and
`year_scaled` combinations present in the sample, with their observed frequency).

**Phylo RE**: by default `marginaleffects` uses `re_formula = NULL`, meaning it conditions
on the fitted phylogenetic random intercepts. This means the AME reflects predictions
for the specific language groups in the sample, not a generic new society.

**In short**: "for the societies in our sample, what is the average change in predicted
probability when violent_external switches from 0 to 1?"

---

### 2. Grand mean AME — section 6 (`m4_grand_ame`)

**How**: `epred_draws()` with `re_formula = NA`, `year_scaled = 0`, expanded over all
unique `world_region` values, then averaged over regions within each posterior draw.

**What it averages over**: all world regions equally (unweighted average), at average year.

**Phylo RE**: `re_formula = NA` strips the phylogenetic random intercept entirely.
Predictions are purely from fixed effects — the population-level trend, not tied to
any specific language group.

**In short**: "at average year, ignoring which language family a society belongs to,
what is the expected change in probability across world regions?"

---

### 3. Empirical AME — section 6 (`m4_empirical_ame`)

**How**: same as grand mean but `newdata` is a random sample of actual observed
`(world_region, year_scaled)` rows from the data (up to n=300 per marker).

**What it averages over**: the observed joint distribution of region and year in the sample.

**Phylo RE**: `re_formula = NA` (same as grand mean).

**In short**: "at the actual distribution of time periods and regions observed in this
marker's data, what is the expected change in probability?"

---

### Key differences to check

| | averages over | phylo RE | year |
|---|---|---|---|
| `avg_slopes` | observed rows | included | observed |
| Grand mean AME | regions equally | excluded | fixed at 0 |
| Empirical AME | observed region × year | excluded | observed |

**Things to verify**:

1. Do the three AME estimates agree? If yes, the effect is robust to these modelling choices.
   If the `avg_slopes` AME (with phylo RE) differs from the grand mean AME (without), this
   tells you something about how much the phylogenetic clustering is inflating or deflating
   the estimated effect.

2. The grand mean and empirical AME use `re_formula = NA`. This is appropriate for asking
   "population-level" questions, but note it is a stronger extrapolation — it asks what would
   happen for a generic society, not specifically for the language groups we observed.

3. `year_scaled = 0` corresponds to the mean year in the sample. Check what actual calendar
   year this maps to for each marker (it will differ slightly since each marker has a different
   set of entries). If the year effect is small (which the model suggests it is), this choice
   should not matter much.

4. The `avg_slopes` approach computes the AME on the logit scale and then transforms, whereas
   the `epred_draws` approach works directly on the probability scale. Both are valid but can
   give slightly different results due to Jensen's inequality. The `epred_draws` approach is
   arguably more honest as it avoids any ambiguity about the order of averaging and
   transforming.
