---
title: "Main Analysis: Ritual markers ~ violent external conflict"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(pacman)
p_load(brms, ape, tidyverse, here, HDInterval, marginaleffects, loo, ggdist)

set.seed(1342)

REFIT <- FALSE  # set TRUE to refit all models from scratch

markers <- c(
  "circumcision", "dress", "extra_ritual_group_markers",
  "food_taboos", "hair", "ornaments",
  "permanent_scarring", "tattoos_scarification"
)

model_names <- c("fit_base", "fit_phylo", "fit_phylo_repeat", "fit_phylo_year")

marker_labels <- c(
  "circumcision"               = "Circumcision",
  "dress"                      = "Dress",
  "extra_ritual_group_markers" = "Extra-Ritual In-Group Markers",
  "food_taboos"                = "Food Taboos",
  "hair"                       = "Hair",
  "ornaments"                  = "Ornaments",
  "permanent_scarring"         = "Permanent Scarring",
  "tattoos_scarification"      = "Tattoos or Scarification"
)

model_labels <- c(
  "fit_base"         = "Baseline",
  "fit_phylo"        = "Phylogenetic",
  "fit_phylo_repeat" = "Phylogenetic + Repeated",
  "fit_phylo_year"   = "Phylogenetic + Year + Region"
)

# Output directories
for (d in c(
  here("data", "mdl_fits", "fit_base"),
  here("data", "mdl_fits", "fit_phylo"),
  here("data", "mdl_fits", "fit_phylo_repeat"),
  here("data", "mdl_fits", "fit_phylo_year"),
  here("data", "mdl_output_main"),
  here("figures", "summary")
)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

for (m in markers)
  dir.create(here("figures", "diagnostics", m), recursive = TRUE, showWarnings = FALSE)
```

---

## 1. Fit models

Set `REFIT <- TRUE` to refit from scratch; otherwise loads cached `.rds` files.

```{r fit_models, message=FALSE, warning=FALSE}
tree <- read.tree(here("asjp", "raw", "world.tre"))

fits <- list()

for (marker in markers) {
  cat("\n--- Processing:", marker, "---\n")

  data <- read_csv(
    here("data", "phylo_input", paste0(marker, ".csv")),
    show_col_types = FALSE
  ) |> mutate(tip_name_rep = tip_name)

  tips        <- unique(data$tip_name)
  tree_pruned <- keep.tip(tree, tips)
  A           <- vcv.phylo(tree_pruned, corr = TRUE)

  f_base  <- formula(paste(marker, "~ 1 + violent_external"))
  f_phylo <- formula(paste(marker, "~ 1 + violent_external + (1 | gr(tip_name, cov = A))"))
  f_rep   <- formula(paste(marker, "~ 1 + violent_external + (1 | gr(tip_name, cov = A)) + (1 | tip_name_rep)"))
  f_year  <- formula(paste(marker, "~ 1 + violent_external + year_scaled + world_region + (1 | gr(tip_name, cov = A))"))

  prior_base  <- c(prior(normal(0, 1.5), class = Intercept), prior(normal(0, 1), class = b))
  prior_phylo <- c(prior_base, prior(exponential(1), class = sd))

  shared <- list(family = bernoulli(link = logit),
                 chains = 4, iter = 4000, warmup = 2000, cores = 4,
                 control = list(adapt_delta = .95, max_treedepth = 15),
                 seed = 1342, silent = 2)

  load_or_fit <- function(model_name, fit_fn) {
    path <- here("data", "mdl_fits", model_name, paste0(marker, ".rds"))
    if (!REFIT && file.exists(path)) {
      readRDS(path)
    } else {
      fit <- fit_fn()
      saveRDS(fit, path)
      fit
    }
  }

  fit_base <- load_or_fit("fit_base", function()
    do.call(brm, c(list(formula = f_base, data = data, prior = prior_base), shared)))

  fit_phylo <- load_or_fit("fit_phylo", function()
    do.call(brm, c(list(formula = f_phylo, data = data, data2 = list(A = A), prior = prior_phylo), shared)))

  fit_phylo_repeat <- load_or_fit("fit_phylo_repeat", function()
    do.call(brm, c(list(formula = f_rep, data = data, data2 = list(A = A), prior = prior_phylo), shared)))

  fit_phylo_year <- load_or_fit("fit_phylo_year", function()
    do.call(brm, c(list(formula = f_year, data = data, data2 = list(A = A), prior = prior_phylo), shared)))

  fits[[marker]] <- list(
    fit_base         = fit_base,
    fit_phylo        = fit_phylo,
    fit_phylo_repeat = fit_phylo_repeat,
    fit_phylo_year   = fit_phylo_year,
    data             = data,
    A                = A
  )
}

# Marker labels with sample sizes
n_labels <- map_chr(markers, function(marker) {
  paste0(marker_labels[marker], " (n=", nrow(fits[[marker]]$data), ")")
}) |> setNames(markers)
```

---

## 2. Extract results

### Fixed effects (log-odds scale)

```{r fixed_effects}
fixed_effects <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    fe <- fixef(fits[[marker]][[mname]], summary = FALSE)
    map_dfr(colnames(fe), function(param) {
      draws <- fe[, param]
      tibble(marker = marker, model = mname, parameter = param,
             median = median(draws), lo95 = hdi(draws)[1], hi95 = hdi(draws)[2])
    })
  })
})

fixed_effects <- fixed_effects |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(fixed_effects, here("data", "mdl_output_main", "coefficients.csv"))
fixed_effects |> filter(parameter == "violent_external")
```

### Hypothesis tests: P(violent_external > 0)

```{r hypothesis_tests}
hypothesis_tests <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    h <- hypothesis(fits[[marker]][[mname]], "violent_external > 0")$hypothesis
    tibble(marker = marker, model = mname,
           estimate   = h$Estimate,
           lo95       = h$CI.Lower,
           hi95       = h$CI.Upper,
           post_prob  = h$Post.Prob,
           evid_ratio = h$Evid.Ratio)
  })
})

hypothesis_tests <- hypothesis_tests |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(hypothesis_tests, here("data", "mdl_output_main", "hypothesis_tests.csv"))
hypothesis_tests
```

### Natural scale: OR and AME

OR = exp(b_violent_external): multiplicative change in odds.

The odds of a marker being present (e.g. tattoos) are X times higher in societies with violent external conflict than in societies without it, holding everything else constant. E.g., OR=2 means double, but applies everywhere one scale (e.g., 1%-->2% or 10%-->20%).

AME = average change in P(outcome = 1) for violent_external 0 → 1, averaged over the data.

Result is in probability units e.g., 0.12 means that, on average across the sample, having violent external conflict is associated with 12 percentage-point higher probability of marker being present. AME accounts for the nonlinearity of the logistic function — the same log-odds coefficient translates to a larger probability shift near P=0.5 and a smaller shift near P=0 or P=1. The AME averages over where each observation actually sits on that curve, so it's more directly interpretable as a real-world effect size. (read more up on this.)

```{r natural_scale}
natural_scale <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    fit  <- fits[[marker]][[mname]]
    data <- fits[[marker]]$data

    # Odds ratio from posterior draws
    draws    <- as_draws_df(fit)$b_violent_external
    or_draws <- exp(draws)

    # Average marginal effect (probability scale)
    ame <- avg_slopes(fit, variables = "violent_external", newdata = data)

    tibble(
      marker     = marker,
      model      = mname,
      or_median  = median(or_draws),
      or_lo95    = hdi(or_draws)[1],
      or_hi95    = hdi(or_draws)[2],
      ame_median = ame$estimate,
      ame_lo95   = ame$conf.low,
      ame_hi95   = ame$conf.high
    )
  })
})

natural_scale <- natural_scale |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(natural_scale, here("data", "mdl_output_main", "natural_scale.csv"))
natural_scale
```

### Phylogenetic signal (λ)

```{r phylo_signal}
phylo_signal <- map_dfr(markers, function(marker) {
  # fit_phylo
  h1 <- hypothesis(
    fits[[marker]]$fit_phylo,
    "sd_tip_name__Intercept^2 / (sd_tip_name__Intercept^2 + 3.289868) = 0",
    class = NULL
  )$hypothesis

  # fit_phylo_repeat
  h2 <- hypothesis(
    fits[[marker]]$fit_phylo_repeat,
    paste("sd_tip_name__Intercept^2 /",
          "(sd_tip_name__Intercept^2 + sd_tip_name_rep__Intercept^2 + 3.289868) = 0"),
    class = NULL
  )$hypothesis

  bind_rows(
    tibble(marker = marker, model = "fit_phylo",        lambda = h1$Estimate, lo95 = h1$CI.Lower, hi95 = h1$CI.Upper),
    tibble(marker = marker, model = "fit_phylo_repeat", lambda = h2$Estimate, lo95 = h2$CI.Lower, hi95 = h2$CI.Upper)
  )
})

phylo_signal <- phylo_signal |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(phylo_signal, here("data", "mdl_output_main", "phylo_signal.csv"))
phylo_signal
```

### Diagnostics (Rhat, ESS)

```{r diagnostics}
diagnostics <- map_dfr(markers, function(marker) {
  map_dfr(model_names, function(mname) {
    s <- summary(fits[[marker]][[mname]])$fixed
    tibble(marker = marker, model = mname, parameter = rownames(s),
           rhat = s[, "Rhat"], ess_bulk = s[, "Bulk_ESS"], ess_tail = s[, "Tail_ESS"])
  })
})

diagnostics <- diagnostics |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(diagnostics, here("data", "mdl_output_main", "diagnostics.csv"))

# Flag any potential issues
diagnostics |> filter(rhat > 1.01 | ess_bulk < 400 | ess_tail < 400)
```

---

## 3. LOO model comparison

```{r loo_comparison, message=FALSE, warning=FALSE}
loo_results <- map_dfr(markers, function(marker) {
  loos <- list(
    fit_base         = loo(fits[[marker]]$fit_base),
    fit_phylo        = loo(fits[[marker]]$fit_phylo),
    fit_phylo_repeat = loo(fits[[marker]]$fit_phylo_repeat),
    fit_phylo_year   = loo(fits[[marker]]$fit_phylo_year)
  )
  comparison <- loo_compare(x = loos)
  as.data.frame(comparison) |>
    rownames_to_column("model") |>
    mutate(marker = marker)
})

loo_results <- loo_results |> mutate(across(where(is.numeric), \(x) round(x, 2)))

write_csv(loo_results, here("data", "mdl_output_main", "loo_comparison.csv"))
loo_results
```

---

## 4. Summary forest plots (all models)

Point + range plots showing all four models per marker.

```{r forest_logodds}
coef_ve <- fixed_effects |>
  filter(parameter == "violent_external") |>
  mutate(
    marker_label = factor(n_labels[marker], levels = n_labels[markers]),
    model_label  = factor(model_labels[model], levels = model_labels)
  )

p <- ggplot(coef_ve, aes(x = median, xmin = lo95, xmax = hi95,
                          y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Coefficient (log-odds)", y = NULL, color = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "All models — posterior median and 95% HDI") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "summary", "forest_logodds.png"), p, width = 10, height = 7)
p
```

```{r forest_or}
ns_plot <- natural_scale |>
  mutate(
    marker_label = factor(n_labels[marker], levels = n_labels[markers]),
    model_label  = factor(model_labels[model], levels = model_labels)
  )

p <- ggplot(ns_plot, aes(x = or_median, xmin = or_lo95, xmax = or_hi95,
                          y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  labs(x = "Odds Ratio (log scale)", y = NULL, color = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "All models — odds ratios") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "summary", "forest_or.png"), p, width = 10, height = 7)
p
```

```{r forest_ame}
p <- ggplot(ns_plot, aes(x = ame_median, xmin = ame_lo95, xmax = ame_hi95,
                          y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Average Marginal Effect (probability scale)", y = NULL, color = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "All models — AME") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "summary", "forest_ame.png"), p, width = 10, height = 7)
p
```

---

## 5. Main model distribution plots (fit_phylo_year)

Posterior distributions for `fit_phylo_year` only. Markers ordered by median effect size.

```{r m4_draws}
# Log-odds posterior draws from fit_phylo_year
draws_logodds <- map_dfr(markers, function(marker) {
  draws <- as_draws_df(fits[[marker]]$fit_phylo_year)$b_violent_external
  tibble(marker = marker, draw = draws)
})

# OR draws
draws_or <- draws_logodds |> mutate(draw = exp(draw))

# AME draws (full posterior distribution)
draws_ame <- map_dfr(markers, function(marker) {
  fit  <- fits[[marker]]$fit_phylo_year
  data <- fits[[marker]]$data
  pd   <- posterior_draws(avg_slopes(fit, variables = "violent_external", newdata = data))
  tibble(marker = marker, draw = pd$draw)
})

# Marker order by median log-odds effect (ascending, so largest is on top)
marker_order <- draws_logodds |>
  group_by(marker) |>
  summarise(med = median(draw), .groups = "drop") |>
  arrange(med) |>
  pull(marker)
```

```{r m4_dist_logodds}
p <- draws_logodds |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Coefficient (log-odds)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "Phylogenetic + Year + Region — posterior distribution") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 11))

ggsave(here("figures", "summary", "m4_dist_logodds.png"), p, width = 9, height = 7)
p
```

```{r m4_dist_or}
p <- draws_or |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  labs(x = "Odds Ratio (log scale)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "Phylogenetic + Year + Region — posterior distribution") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 11))

ggsave(here("figures", "summary", "m4_dist_or.png"), p, width = 9, height = 7)
p
```

```{r m4_dist_ame}
# Use same marker order as log-odds
ame_order <- draws_ame |>
  group_by(marker) |>
  summarise(med = median(draw), .groups = "drop") |>
  arrange(med) |>
  pull(marker)

p <- draws_ame |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[ame_order])) |>
  ggplot(aes(x = draw, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_hdi,
               fill = "#C8839E", color = "black", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Average Marginal Effect (probability scale)", y = NULL,
       title = "Effect of violent external conflict on ritual markers",
       subtitle = "Phylogenetic + Year + Region — posterior distribution") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 11))

ggsave(here("figures", "summary", "m4_dist_ame.png"), p, width = 9, height = 7)
p
```

## 7. Phylogenetic signal plot

```{r phylo_signal_plot}
p <- phylo_signal |>
  mutate(
    marker_label = factor(n_labels[marker], levels = n_labels[markers]),
    model_label  = factor(model_labels[model], levels = model_labels)
  ) |>
  ggplot(aes(x = lambda, xmin = lo95, xmax = hi95, y = marker_label, color = model_label)) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Phylogenetic signal (λ)", y = NULL, color = NULL,
       title = "Phylogenetic signal across markers",
       subtitle = "Posterior median and 95% HDI") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures", "summary", "phylo_signal.png"), p, width = 10, height = 7)
p
```
