---
title: "1_external"
output: html_document
---

Analysis of marker ~ external violent conflict.
First fits baseline models, then adds phylogenetic control.
Fits models for all markers. 

```{r}
library(pacman)
p_load(brms, tidyverse, here, ggdist, ape, ggtext, marginaleffects)
set.seed(12532)

# marginaleffects centres Bayesian estimates on the median by default.
# should not matter a ton, but we use mean throughout in other analyses.
# so just keeping it consistent here.
options(marginaleffects_posterior_center = "mean")

dir.create(here("data","model","external","results"), recursive = TRUE, showWarnings = FALSE)
```

```{r}

# all markers 
markers <- c(
  "circumcision", "dress", "extra_ritual_group_markers", "food_taboos",
  "hair", "ornaments", "permanent_scarring", "tattoos_scarification"
)

# labels for markers (nicer for plots, outputs)
marker_labels <- c(
  circumcision="Circumcision", dress="Dress",
  extra_ritual_group_markers="Extra-Ritual In-Group Markers",
  food_taboos="Food Taboos", hair="Hair", ornaments="Ornaments",
  permanent_scarring="Permanent Scarring", tattoos_scarification="Tattoos/Scarification"
)

# reporting order is that the two top-level markers go first, then the rest.
markers_table_order <- c("permanent_scarring", "extra_ritual_group_markers",
                          "circumcision", "dress", "food_taboos",
                          "hair", "ornaments", "tattoos_scarification")

# for formatting outputs to clearly communicate structure.
marker_level <- c(
  permanent_scarring = "parent",
  extra_ritual_group_markers = "parent",
  circumcision = "sub",
  dress = "sub",
  food_taboos = "sub",
  hair = "sub",
  ornaments = "sub",
  tattoos_scarification = "sub"
)

# priors 
priors <- c(
  prior(normal(0, 1.5), class = Intercept), 
  prior(normal(0, 1),   class = b), 
  prior(exponential(1), class = sd) 
)

# list to hold fits 
fits <- list()
for (m in markers) {
  d <- read_csv(here("data","model","external","input", paste0(m, ".csv")),
                show_col_types = FALSE) |>
    
    # technically could drop the world_region line given Africa first 
    # alphabetically, and only used for tables (not main eff.) which is a bit misleading.
    mutate(
      y = as.integer(.data[[m]]), 
      world_region = fct_relevel(factor(world_region), "Africa")) 

  f <- bf(as.formula(paste(m, "~ 1 + violent_external + year_scaled + (1 | world_region)")))

  fit <- brm(f, data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

  fits[[m]] <- list(fit = fit, data = d, n = nobs(fit))
}

```

Convergence

```{r diagnostics_baseline}

# per-marker check convergence
diag_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  fixed <- summary(fit)$fixed
  sds <- bind_rows(map(summary(fit)$random, as.data.frame))

  tibble(
    Marker = marker_labels[m],
    N = fits[[m]]$n,
    divergences = sum(subset(nuts_params(fit), Parameter=="divergent__")$Value),
    max_rhat = round(max(rhat(fit), na.rm=TRUE), 3),
    min_bulk_ess = round(min(c(fixed[,"Bulk_ESS"], sds[,"Bulk_ESS"])), 0),
    min_tail_ess = round(min(c(fixed[,"Tail_ESS"], sds[,"Tail_ESS"])), 0)
  )
})
diag_tbl
write_csv(diag_tbl, here("data","model","external","results","diagnostics_baseline.csv"))

```

Posterior predictive check: observed prevalence against its posterior predictive
interval, split by conflict status. 

```{r ppc_baseline}

ppc <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  d <- fits[[m]]$data
  yrep <- posterior_predict(fit, ndraws = 1000)
  conflict <- ifelse(d$violent_external == 1, "Yes", "No")

  map_dfr(c("No", "Yes"), function(g) {
    idx <- which(conflict == g)
    prop <- rowMeans(yrep[, idx, drop = FALSE])
    tibble(
      Marker = unname(marker_labels[m]),
      conflict = g,
      observed = mean(d[[m]][idx]),
      lo = as.numeric(quantile(prop, 0.025)),
      hi = as.numeric(quantile(prop, 0.975))
    )
  })
}) |>
  mutate(Marker = factor(Marker, levels = rev(unname(marker_labels[markers_table_order]))))

ggplot(ppc, aes(y = Marker, colour = conflict)) +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 3, alpha = 0.4,
                 position = position_dodge(width = 0.5)) +
  geom_point(aes(x = observed), size = 2,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c(No = "#8FBF9F", Yes = "#D98B72")) +
  labs(x = "Marker prevalence", y = NULL, colour = "External conflict",
       subtitle = "Bars: 95% posterior predictive interval. Points: observed proportion.") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

```

Fixed effects table

```{r fixed_effects_baseline}

# one row per marker; each term formatted as "estimate [lo, hi]" to go straight into paper.
fmt <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)

fixed_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  fx <- fixef(fit) # rows: Intercept, violent_external, year_scaled

  tibble(
    Marker = marker_labels[m],
    N = fits[[m]]$n,
    Intercept = fmt(
      fx["Intercept","Estimate"],
      fx["Intercept","Q2.5"],
      fx["Intercept","Q97.5"]),
    violent_external = fmt(
      fx["violent_external","Estimate"],
      fx["violent_external","Q2.5"],
      fx["violent_external","Q97.5"]),
    year_scaled = fmt(
      fx["year_scaled","Estimate"],
      fx["year_scaled","Q2.5"],
      fx["year_scaled","Q97.5"])
  )
})

fixed_tbl
write_csv(fixed_tbl, here("data","model","external","results","fixed_effects_baseline.csv"))

```

Table for random effects (world region.)
To really compare whether world region is predictive, we probably would need to fit also without and do model comparison.
But this is not the main objective, and pretty clear that for some markers there is regional heterogeneity.

```{r}

sd_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  rf <- summary(fit)$random$world_region  # row: sd(Intercept)

  tibble(
    Marker = marker_labels[m],
    N = fits[[m]]$n,
    sd_world_region = fmt(
      rf["sd(Intercept)","Estimate"],
      rf["sd(Intercept)","l-95% CI"],
      rf["sd(Intercept)","u-95% CI"])
  )
})

sd_tbl
write_csv(sd_tbl, here("data","model","external","results","random_effects_baseline.csv"))

```

Per-region breakdown: raw (unpooled) rate vs. model-estimated (partially pooled) rate,
to see which regions drive a marker's regional heterogeneity.

```{r region_effects_baseline}

region_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  d <- fits[[m]]$data

  raw <- d |> group_by(world_region) |>
    summarise(n = n(), raw_rate = mean(y), .groups = "drop")

  # predict at each entry's own conflict value and start year so the estimate is
  # conditioned like raw_rate; average within region per draw, then across draws
  ep <- posterior_epred(fit, newdata = d, re_formula = ~ (1 | world_region))
  regions <- sort(unique(as.character(d$world_region)))

  model_est <- map_dfr(regions, function(r) {
    idx <- which(as.character(d$world_region) == r)
    p <- rowMeans(ep[, idx, drop = FALSE])
    tibble(
      world_region = r,
      model_estimate = mean(p),
      ci_lo = as.numeric(quantile(p, 0.025)),
      ci_hi = as.numeric(quantile(p, 0.975)))
  })

  raw |>
    mutate(world_region = as.character(world_region)) |>
    left_join(model_est, by = "world_region") |>
    mutate(Marker = marker_labels[m], .before = 1)
})

region_tbl <- region_tbl |>
  mutate(across(c(raw_rate, model_estimate, ci_lo, ci_hi), \(x) round(x, 2)))

region_tbl
write_csv(region_tbl, here("data","model","external","results","region_effects_baseline.csv"))

```

Hypothesis test + AME.

```{r}

# G-computation over the observed covariate distribution; draws are reused by the plots below.
# avg_predictions rather than avg_comparisons: differencing the two levels gives the same
# contrast. But we need this for the base-rates plot
ame_all_baseline <- map_dfr(markers_table_order, function(m) {
  avg_predictions(fits[[m]]$fit, variables = list(violent_external = c(0, 1)), re_formula = NA) |>
    posterior_draws() |>
    select(drawid, violent_external, draw) |>
    pivot_wider(names_from = violent_external, values_from = draw, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0, marker = m)
})

results_tbl <- map_dfr(markers_table_order, function(m) {
  h <- hypothesis(fits[[m]]$fit, "violent_external > 0", alpha = 0.025)$hypothesis
  ame <- ame_all_baseline$ame[ame_all_baseline$marker == m]

  tibble(
    Marker = marker_labels[m],
    N = fits[[m]]$n,
    beta = h$Estimate,
    ci_lo = h$CI.Lower,
    ci_hi = h$CI.Upper,
    post_prob = h$Post.Prob,
    AME = mean(ame),
    AME_lo = as.numeric(quantile(ame, 0.025)),
    AME_hi = as.numeric(quantile(ame, 0.975))
  )
}) |> mutate(across(where(is.numeric), \(x) round(x, 2)))

results_tbl
write_csv(results_tbl, here("data","model","external","results","hypothesis_ame_baseline.csv"))

```

Plot estimated base rates: baseline model's predicted probability under violent_external=0 vs.
=1, per marker using the AME above.

```{r}

marker_order <- ame_all_baseline |> group_by(marker) |>
  summarise(md = median(ame)) |> arrange(md) |> pull(marker)

# separate ordering for the base-rates plot: sorted by prevalence under cond1.
marker_order_prevalence <- ame_all_baseline |> group_by(marker) |>
  summarise(md = median(cond1)) |> arrange(md) |> pull(marker)

# parent markers (Permanent Scarring, Extra-Ritual In-Group Markers) get **bold** markdown
# around their label, rendered via ggtext::element_markdown() on the axis text below.
n_labels <- map_chr(markers_table_order, \(m) {
  lbl <- paste0(marker_labels[m], " (n=", fits[[m]]$n, ")")
  if (marker_level[m] == "parent") paste0("**", lbl, "**") else lbl
}) |> setNames(markers_table_order)

p_base_rates <- ame_all_baseline |>
  pivot_longer(c(cond0, cond1), names_to = "condition", values_to = "p") |>
  mutate(condition = fct_recode(condition,
                                 "No violent conflict" = "cond0",
                                 "Violent conflict" = "cond1"),
         marker_label = factor(n_labels[marker], levels = n_labels[marker_order_prevalence])) |>
  ggplot(aes(x = p, y = marker_label, fill = condition)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               position = position_dodge(width = 0.3), scale = 1.5,
               color = "black", alpha = 0.7) +
  scale_fill_manual(values = c("No violent conflict" = "#8FBF9F",
                                "Violent conflict" = "#D98B72")) +
  labs(x = "Estimated prevalence", y = NULL, fill = NULL) +
  theme_bw(base_size = 16) +               # <- bumps ALL text at once
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16),
    axis.title.x = element_text(size = 17),
    axis.text.y = element_markdown(size = 15),
    axis.text.x = element_text(size = 14),
    legend.position = "top",
    plot.margin = margin(12, 16, 12, 16)
  )

ggsave(here("data","model","external","results","external_base_rates_baseline.pdf"),
       p_base_rates, width = 9, height = 7, device = "pdf")
p_base_rates

```


## PHYLO ##

```{r}

tree <- read.tree(here("data","jaeger2018","world.tre"))   # load ONCE, prune per marker

```

Fit models.

```{r fit_phylo}
fits_phylo <- list()
for (m in markers) {
  d <- read_csv(here("data","model","external","input", paste0(m,".csv")),
                show_col_types = FALSE) |>
    mutate(y = as.integer(.data[[m]]),
           world_region = fct_relevel(factor(world_region), "Africa"))

  # phylo sample = mapped entries only; duplicate tip col for the two effects
  d_phylo <- d |> filter(!is.na(tip_name)) |> mutate(phylo = tip_name)

  # per-marker tree prune + correlation matrix
  tips <- unique(d_phylo$tip_name)
  stopifnot(all(tips %in% tree$tip.label)) # guard 
  tree_pruned <- keep.tip(tree, tips)
  A <- vcv.phylo(tree_pruned, corr = TRUE)
  stopifnot(all(d_phylo$phylo %in% rownames(A)))

  f <- bf(as.formula(paste(
    m, "~ 1 + violent_external + year_scaled +",
    "(1 | world_region) + (1 | gr(phylo, cov = A)) + (1 | tip_name)"
  )))

  fit <- brm(f, data = d_phylo, data2 = list(A = A),
             family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99, max_treedepth = 12),
             seed = 52312, backend = "cmdstanr", silent = 2)

  fits_phylo[[m]] <- list(fit = fit, data = d_phylo, n = nobs(fit),
                    n_tips = length(tips))
}
```

Per-marker convergence.

```{r diagnostics_phylo}

# per-marker convergence sweep — CATCH bad fits before trusting results
diag_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  fixed <- summary(fit)$fixed
  sds <- bind_rows(map(summary(fit)$random, as.data.frame))

  tibble(
    Marker = marker_labels[m],
    N = fits_phylo[[m]]$n,
    n_tips = fits_phylo[[m]]$n_tips,
    divergences = sum(subset(nuts_params(fit), Parameter=="divergent__")$Value),
    max_rhat = round(max(rhat(fit), na.rm=TRUE), 3),
    min_bulk_ess = round(min(c(fixed[,"Bulk_ESS"], sds[,"Bulk_ESS"])), 0),
    min_tail_ess = round(min(c(fixed[,"Tail_ESS"], sds[,"Tail_ESS"])), 0)
  )
})
diag_tbl
write_csv(diag_tbl, here("data","model","external","results","diagnostics_phylo.csv"))

```

Posterior predictive check: observed prevalence against its posterior predictive
interval, split by conflict status. A density overlay would be uninformative for a
0/1 outcome.

```{r ppc_phylo}

ppc <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  d <- fits_phylo[[m]]$data
  yrep <- posterior_predict(fit, ndraws = 1000)
  conflict <- ifelse(d$violent_external == 1, "Yes", "No")

  map_dfr(c("No", "Yes"), function(g) {
    idx <- which(conflict == g)
    prop <- rowMeans(yrep[, idx, drop = FALSE])
    tibble(
      Marker = unname(marker_labels[m]),
      conflict = g,
      observed = mean(d[[m]][idx]),
      lo = as.numeric(quantile(prop, 0.025)),
      hi = as.numeric(quantile(prop, 0.975))
    )
  })
}) |>
  mutate(Marker = factor(Marker, levels = rev(unname(marker_labels[markers_table_order]))))

ggplot(ppc, aes(y = Marker, colour = conflict)) +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 3, alpha = 0.4,
                 position = position_dodge(width = 0.5)) +
  geom_point(aes(x = observed), size = 2,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c(No = "#8FBF9F", Yes = "#D98B72")) +
  labs(x = "Marker prevalence", y = NULL, colour = "External conflict",
       subtitle = "Bars: 95% posterior predictive interval. Points: observed proportion.") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

```

Fixed effects table

```{r fixed_effects_phylo}

# one row per marker; each term formatted as "estimate [lo, hi]" — same shape as fixed_effects_baseline
fixed_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  fx <- fixef(fit) # rows: Intercept, violent_external, year_scaled

  tibble(
    Marker = marker_labels[m],
    N = fits_phylo[[m]]$n,
    Intercept = fmt(
      fx["Intercept","Estimate"],
      fx["Intercept","Q2.5"],
      fx["Intercept","Q97.5"]),
    violent_external = fmt(
      fx["violent_external","Estimate"],
      fx["violent_external","Q2.5"],
      fx["violent_external","Q97.5"]),
    year_scaled = fmt(
      fx["year_scaled","Estimate"],
      fx["year_scaled","Q2.5"],
      fx["year_scaled","Q97.5"])
  )
})

fixed_tbl
write_csv(fixed_tbl, here("data","model","external","results","fixed_effects_phylo.csv"))

```

Random effects table: one row per marker, sd(Intercept) for world_region, phylo (covariance-
structured tip effect), and tip_name (iid tip effect) — same "est [lo, hi]" format as baseline.

```{r random_effects_phylo}

sd_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  rf_region <- summary(fit)$random$world_region
  rf_phylo <- summary(fit)$random$phylo
  rf_tip <- summary(fit)$random$tip_name

  tibble(
    Marker = marker_labels[m],
    N = fits_phylo[[m]]$n,
    sd_world_region = fmt(
      rf_region["sd(Intercept)","Estimate"],
      rf_region["sd(Intercept)","l-95% CI"],
      rf_region["sd(Intercept)","u-95% CI"]),
    sd_phylo = fmt(
      rf_phylo["sd(Intercept)","Estimate"],
      rf_phylo["sd(Intercept)","l-95% CI"],
      rf_phylo["sd(Intercept)","u-95% CI"]),
    sd_tip_name = fmt(
      rf_tip["sd(Intercept)","Estimate"],
      rf_tip["sd(Intercept)","l-95% CI"],
      rf_tip["sd(Intercept)","u-95% CI"])
  )
})

sd_tbl
write_csv(sd_tbl, here("data","model","external","results","random_effects_phylo.csv"))

```

Phylogenetic signal: proportion of tip-level variance that is phylogenetically structured
(sd_phylo, tree-respecting) vs. idiosyncratic per-tip (sd_tip_name). Computed per posterior
draw so the ratio itself carries uncertainty, not just from the two point estimates.

This is a variance ratio (phylogenetic heritability), NOT Pagel's lambda -- we never
estimate a tree-transformation parameter. Bernoulli has no residual sd, so sd_tip_name
plays the role Gaussian sigma would; it is identified only because entries outnumber tips.
The denominator is tip-level variance only: sd_world_region and the logistic link's
pi^2/3 are both excluded, so this is not a share of total latent variance. Tables label
the column with the formula for exactly this reason.

```{r phylo_signal}

phylo_signal_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  draws <- as_draws_df(fit, variable = c("sd_phylo__Intercept", "sd_tip_name__Intercept"))
  h2 <- draws$sd_phylo__Intercept^2 /
    (draws$sd_phylo__Intercept^2 + draws$sd_tip_name__Intercept^2)

  tibble(
    Marker = marker_labels[m],
    N = fits_phylo[[m]]$n,
    phylo_signal = fmt(
      mean(h2),
      quantile(h2, 0.025),
      quantile(h2, 0.975))
  )
})

phylo_signal_tbl
write_csv(phylo_signal_tbl, here("data","model","external","results","phylo_signal.csv"))

```

AME + hypothesis table

```{r results_table}

# re_formula = NA excludes ALL random effects (region, phylo, tip); draws reused below
ame_all_phylo <- map_dfr(markers_table_order, function(m) {
  avg_predictions(fits_phylo[[m]]$fit, variables = list(violent_external = c(0, 1)), 
                  re_formula = NA) |>
    posterior_draws() |>
    select(drawid, violent_external, draw) |>
    pivot_wider(names_from = violent_external, values_from = draw, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0, marker = m)
})

results_tbl <- map_dfr(markers_table_order, function(m) {
  h <- hypothesis(fits_phylo[[m]]$fit, "violent_external > 0", alpha = 0.025)$hypothesis
  ame <- ame_all_phylo$ame[ame_all_phylo$marker == m]

  tibble(
    Marker = marker_labels[m],
    N = fits_phylo[[m]]$n,
    beta = h$Estimate,
    ci_lo = h$CI.Lower,
    ci_hi = h$CI.Upper,
    post_prob = h$Post.Prob,
    AME = mean(ame),
    AME_lo = as.numeric(quantile(ame, 0.025)),
    AME_hi = as.numeric(quantile(ame, 0.975))
  )
}) |> mutate(across(where(is.numeric), \(x) round(x, 2)))

results_tbl
write_csv(results_tbl, here("data","model","external","results","hypothesis_ame_phylo.csv"))

```

Plot AME distributions: baseline vs. phylo, to see how much the violent_external effect
shifts once the two tip-level random effects (phylogenetic and non-phylogenetic) are added.

```{r}

ame_compare <- bind_rows(
  ame_all_baseline |> mutate(model = "Baseline"),
  ame_all_phylo    |> mutate(model = "Phylogenetic")
) |> mutate(model = fct_relevel(model, "Baseline", "Phylogenetic"))

# marker order + labels from the baseline plot above, with N per model (fits/fits_phylo are
# separate objects now, so both are still in memory here — no need to read baseline N back from disk)
# parent markers get **bold** markdown around their label, rendered via
# ggtext::element_markdown() on the axis text below — same trick as p_base_rates.
n_labels_compare <- map_chr(markers, \(m) {
  lbl <- paste0(marker_labels[m], " (n=", fits[[m]]$n, "/", fits_phylo[[m]]$n, ")")
  if (marker_level[m] == "parent") paste0("**", lbl, "**") else lbl
}) |> setNames(markers)

p_ame_compare <- ame_compare |>
  mutate(marker_label = factor(n_labels_compare[marker], levels = n_labels_compare[marker_order])) |>
  ggplot(aes(x = ame, y = marker_label, fill = model)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               position = position_dodge(width = 0.6), scale = 1.1,
               color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c(Baseline = "#C8839E", Phylogenetic = "#6B9AC4")) +
  labs(x = "Average marginal effect (probability)", y = NULL, fill = NULL) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16),
    axis.title.x = element_text(size = 17),
    axis.text.y = element_markdown(size = 15),
    axis.text.x = element_text(size = 14),
    legend.position = "top",
    plot.margin = margin(12, 16, 12, 16)
  )

ggsave(here("data","model","external","results","external_ame_baseline_vs_phylo.pdf"),
       p_ame_compare, width = 9, height = 7, device = "pdf")
p_ame_compare

```

## DEEP DIVE: TATTOOS/SCARIFICATION ##

Per-tip breakdown: raw (unpooled) rate vs. model-estimated (partially pooled) rate, grouped by
language family (first dot-separated segment of the tip name), built the same way as
region_effects_baseline. The estimate is predicted at each entry's observed conflict value and
start year, including the phylogenetic and tip-level random effects, and averaged within tip
per draw. Predictions are used rather than the raw random-effect offsets, since for a
phylogenetically correlated effect the offsets are not anchored on their own.

```{r tip_effects_tattoos_scarification}

m <- "tattoos_scarification"
fit <- fits_phylo[[m]]$fit
d <- fits_phylo[[m]]$data

# one column per entry; reused by the family table below. `A` in re_formula is resolved
# from the covariance stored in the fit (fit$data2), not from the workspace
ep <- posterior_epred(fit, newdata = d, re_formula = ~ (1 | gr(phylo, cov = A)) + (1 | tip_name))
family <- str_split_i(d$tip_name, "\\.", 1)

tip_tbl <- map_dfr(sort(unique(d$tip_name)), function(tp) {
  idx <- which(d$tip_name == tp)
  p <- rowMeans(ep[, idx, drop = FALSE])
  tibble(
    family = family[idx[1]],
    tip_name = tp,
    n = length(idx),
    raw_rate = mean(d$y[idx]),
    model_estimate = mean(p),
    ci_lo = as.numeric(quantile(p, 0.025)),
    ci_hi = as.numeric(quantile(p, 0.975))
  )
}) |>
  arrange(family, desc(model_estimate)) |>
  mutate(across(c(raw_rate, model_estimate, ci_lo, ci_hi), \(x) round(x, 2)))

tip_tbl

```

Aggregated to language family, the level we report at. Both rates are entry-weighted: the
estimate averages the same entry-level predictions within family per draw, which also gives
its credible interval.

```{r family_effects_tattoos_scarification}

family_tbl <- map_dfr(sort(unique(family)), function(f) {
  idx <- which(family == f)
  p <- rowMeans(ep[, idx, drop = FALSE])
  tibble(
    family = f,
    total_n = length(idx),
    n_tips = n_distinct(d$tip_name[idx]),
    raw_rate = mean(d$y[idx]),
    model_estimate = mean(p),
    ci_lo = as.numeric(quantile(p, 0.025)),
    ci_hi = as.numeric(quantile(p, 0.975))
  )
}) |>
  arrange(desc(total_n)) |>
  mutate(across(c(raw_rate, model_estimate, ci_lo, ci_hi), \(x) round(x, 2)))

family_tbl
write_csv(family_tbl, here("data","model","external","results","family_effects_tattoos_scarification.csv"))

```

Versions

```{r versions}

version_pkgs <- c("brms", "tidyverse", "here", "ggdist", "ape", "ggtext",
                  "marginaleffects", "cmdstanr", "posterior")

versions_tbl <- tibble(
  component = c("R", version_pkgs, "cmdstan"),
  version = c(paste(R.version$major, R.version$minor, sep = "."),
              map_chr(version_pkgs, \(p) as.character(packageVersion(p))),
              as.character(cmdstanr::cmdstan_version()))
)

versions_tbl
write_csv(versions_tbl, here("data","model","external","results","versions.csv"))

```

