---
title: "4_external_noeHRAF"
output: html_document
---

Analysis of marker ~ external violent conflict, excluding eHRAF entries (robustness
check on the main external-conflict analysis in 1_external.Rmd).
First fits baseline models, then adds phylogenetic control.
Fits models for all markers.

```{r}
library(pacman)
p_load(brms, tidyverse, here, tidybayes, ggdist, HDInterval, ape, ggtext)
set.seed(12532)

dir.create(here("data","model","external_noeHRAF","results"), recursive = TRUE, showWarnings = FALSE)
```

```{r}

# all markers 
markers <- c(
  "circumcision", "dress", "extra_ritual_group_markers", "food_taboos",
  "hair", "ornaments", "permanent_scarring", "tattoos_scarification"
)

# labels for markers (nicer for plots, outputs)
marker_labels <- c(
  circumcision="Circumcision", dress="Dress",
  extra_ritual_group_markers="Extra-Ritual In-Group Markers",
  food_taboos="Food Taboos", hair="Hair", ornaments="Ornaments",
  permanent_scarring="Permanent Scarring", tattoos_scarification="Tattoos/Scarification"
)

# fixed reporting order for all per-marker tables/plots: the two DRH top-level categories
# first (Permanent Scarring, then Extra-Ritual In-Group Markers, whose sub-questions are the
# other six), then the six sub-markers alphabetically. Single source of truth, reused for
# table row order and for de-emphasizing (lower alpha) the six sub-markers in plots.
markers_table_order <- c("permanent_scarring", "extra_ritual_group_markers",
                          "circumcision", "dress", "food_taboos",
                          "hair", "ornaments", "tattoos_scarification")

marker_level <- c(
  permanent_scarring          = "parent",
  extra_ritual_group_markers  = "parent",
  circumcision                = "sub",
  dress                       = "sub",
  food_taboos                 = "sub",
  hair                        = "sub",
  ornaments                   = "sub",
  tattoos_scarification       = "sub"
)

# priors 
priors <- c(
  prior(normal(0, 1.5), class = Intercept), # wide enough, centering a bit skewed since the empirical rate <50% but fine.
  prior(normal(0, 1),   class = b), # also wide enough (95% includes up to 2 logit shift.)
  prior(exponential(1), class = sd) # could consider exponential(2) because our regions are so loopsided (to regularize more)
)

# list to hold fits 
fits <- list()
for (m in markers) {
  d <- read_csv(here("data","model","external_noeHRAF","input", paste0(m, ".csv")),
                show_col_types = FALSE) |>
    mutate(
      y = as.integer(.data[[m]]), 
      world_region = fct_relevel(factor(world_region), "Africa")) 

  f <- bf(as.formula(paste(m, "~ 1 + violent_external + year_scaled + (1 | world_region)")))

  fit <- brm(f, data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 4000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

  fits[[m]] <- list(fit = fit, data = d, n = nrow(d))
}

```


Convergence

```{r diagnostics_baseline}

# per-marker check convergence
diag_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  tibble(
    Marker      = marker_labels[m],
    N           = fits[[m]]$n,
    divergences   = sum(subset(nuts_params(fit), Parameter=="divergent__")$Value),
    max_rhat      = round(max(rhat(fit), na.rm=TRUE), 3),
    min_bulk_ess  = round(min(summary(fit)$fixed[,"Bulk_ESS"]), 0),
    min_tail_ess  = round(min(summary(fit)$fixed[,"Tail_ESS"]), 0)
  )
})
diag_tbl
write_csv(diag_tbl, here("data","model","external_noeHRAF","results","diagnostics_baseline.csv"))

```

Fixed effects table

```{r fixed_effects_baseline}

# one row per marker; each term formatted as "estimate [lo, hi]" to go straight into paper.
fmt <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)

fixed_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  fx  <- fixef(fit) # rows: Intercept, violent_external, year_scaled

  tibble(
    Marker           = marker_labels[m],
    N                = fits[[m]]$n,
    Intercept        = fmt(fx["Intercept","Estimate"], fx["Intercept","Q2.5"], fx["Intercept","Q97.5"]),
    violent_external = fmt(fx["violent_external","Estimate"], fx["violent_external","Q2.5"], fx["violent_external","Q97.5"]),
    year_scaled      = fmt(fx["year_scaled","Estimate"], fx["year_scaled","Q2.5"], fx["year_scaled","Q97.5"])
  )
})

fixed_tbl
write_csv(fixed_tbl, here("data","model","external_noeHRAF","results","fixed_effects_baseline.csv"))

```

Table for random effects (world region.)
To really compare whether world region is predictive, we probably would need to fit also without and do model comparison.
But this is not the main objective, and pretty clear that for some markers there is regional heterogeneity.

```{r}

sd_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  rf  <- summary(fit)$random$world_region  # row: sd(Intercept)

  tibble(
    Marker          = marker_labels[m],
    N               = fits[[m]]$n,
    sd_world_region = fmt(rf["sd(Intercept)","Estimate"],
                           rf["sd(Intercept)","l-95% CI"],
                           rf["sd(Intercept)","u-95% CI"])
  )
})

sd_tbl
write_csv(sd_tbl, here("data","model","external_noeHRAF","results","random_effects_baseline.csv"))

```

Per-region breakdown: raw (unpooled) rate vs. model-estimated (partially pooled) rate,
to see which regions drive a marker's regional heterogeneity.

```{r region_effects_baseline}

region_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  d   <- fits[[m]]$data

  raw <- d |> group_by(world_region) |>
    summarise(n = n(), raw_rate = mean(y), .groups = "drop")

  nd <- tibble(world_region = unique(d$world_region), violent_external = 0, year_scaled = 0)
  model_est <- fit |> epred_draws(newdata = nd, re_formula = ~ (1 | world_region)) |> 
    ungroup() |>
    group_by(world_region) |>
    summarise(model_estimate = mean(.epred),
              ci_lo = quantile(.epred, 0.025),
              ci_hi = quantile(.epred, 0.975), .groups = "drop")

  raw |> left_join(model_est, by = "world_region") |>
    mutate(Marker = marker_labels[m], .before = 1)
})

region_tbl <- region_tbl |>
  mutate(across(c(raw_rate, model_estimate, ci_lo, ci_hi), \(x) round(x, 2)))

region_tbl
write_csv(region_tbl, here("data","model","external_noeHRAF","results","region_effects_baseline.csv"))

```



Hypothesis test + AME.

```{r}

results_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit
  d   <- fits[[m]]$data

  h <- hypothesis(fit, "violent_external > 0", alpha = 0.025)$hypothesis

  nd <- expand_grid(violent_external = c(0,1), year_scaled = 0,
                    world_region = unique(d$world_region))
  ame <- fit |> epred_draws(newdata = nd, re_formula = NA) |> ungroup() |>
    group_by(.draw, violent_external) |>
    summarise(p = mean(.epred), .groups = "drop") |>
    pivot_wider(names_from = violent_external, values_from = p, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0) |> pull(ame)

  tibble(
    Marker = marker_labels[m],
    N = fits[[m]]$n,
    beta = h$Estimate,
    ci_lo = h$CI.Lower,
    ci_hi = h$CI.Upper,
    post_prob = h$Post.Prob,
    AME = mean(ame)
  )
}) |> mutate(across(where(is.numeric), \(x) round(x, 2)))

results_tbl
write_csv(results_tbl, here("data","model","external_noeHRAF","results","hypothesis_ame_baseline.csv"))

```

Plot estimated base rates: baseline model's predicted probability under violent_external=0 vs.
=1, per marker — distinct from the AME/beta plots below, which only show the difference between
these two distributions, not their absolute level. Built from the same draws as the AME
calculation above (cond0/cond1, before collapsing to the difference), so no new computation is
needed. Sorted by effect size, same alpha convention as the tables (parents at full opacity).

```{r}

# AME draws for ALL markers (kept as draws for the plot); "_baseline" suffix kept for
# clarity alongside ame_all_phylo, used in the baseline-vs-phylo comparison plot below.
ame_all_baseline <- map_dfr(markers_table_order, function(m) {
  fit <- fits[[m]]$fit; d <- fits[[m]]$data
  nd <- expand_grid(violent_external = c(0,1), year_scaled = 0,
                    world_region = unique(d$world_region))
  fit |> epred_draws(newdata = nd, re_formula = NA) |> ungroup() |>
    group_by(.draw, violent_external) |>
    summarise(p = mean(.epred), .groups = "drop") |>
    pivot_wider(names_from = violent_external, values_from = p, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0, marker = m)
})

marker_order <- ame_all_baseline |> group_by(marker) |>
  summarise(md = median(ame)) |> arrange(md) |> pull(marker)

# separate ordering for the base-rates plot: sorted by prevalence under cond1 (the focal
# condition), not by AME. AME can be negative for some markers and positive for others, so
# sorting by the *difference* doesn't give a visually clean prevalence ordering — sorting by
# one condition's actual level does. marker_order (AME-based) stays as-is for the AME
# comparison plot below, where sorting by effect size is what's wanted.
marker_order_prevalence <- ame_all_baseline |> group_by(marker) |>
  summarise(md = median(cond1)) |> arrange(md) |> pull(marker)

# parent markers (Permanent Scarring, Extra-Ritual In-Group Markers) get **bold** markdown
# around their label, rendered via ggtext::element_markdown() on the axis text below.
n_labels <- map_chr(markers_table_order, \(m) {
  lbl <- paste0(marker_labels[m], " (n=", fits[[m]]$n, ")")
  if (marker_level[m] == "parent") paste0("**", lbl, "**") else lbl
}) |> setNames(markers_table_order)

p_base_rates <- ame_all_baseline |>
  pivot_longer(c(cond0, cond1), names_to = "condition", values_to = "p") |>
  mutate(condition = fct_recode(condition,
                                 "No violent conflict" = "cond0",
                                 "Violent conflict"    = "cond1"),
         marker_label = factor(n_labels[marker], levels = n_labels[marker_order_prevalence])) |>
  ggplot(aes(x = p, y = marker_label, fill = condition)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               position = position_dodge(width = 0.3), scale = 1.5,
               color = "black", alpha = 0.7) +
  scale_fill_manual(values = c("No violent conflict" = "#8FBF9F",
                                "Violent conflict"    = "#D98B72")) +
  labs(x = "Estimated prevalence", y = NULL, fill = NULL) +
  theme_bw(base_size = 16) +               # <- bumps ALL text at once
  theme(
    plot.title    = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16),
    axis.title.x  = element_text(size = 17),
    axis.text.y   = element_markdown(size = 15),
    axis.text.x   = element_text(size = 14),
    legend.position = "top",
    plot.margin   = margin(12, 16, 12, 16)
  )

ggsave(here("data","model","external_noeHRAF","results","external_noeHRAF_base_rates_baseline.pdf"),
       p_base_rates, width = 9, height = 7, device = "pdf")
p_base_rates

```


## PHYLO ##

```{r}

tree <- read.tree(here("data","jaeger2018","world.tre"))   # load ONCE, prune per marker

```

Fit models.

```{r fit_phylo}
fits_phylo <- list()
for (m in markers) {
  d <- read_csv(here("data","model","external_noeHRAF","input", paste0(m,".csv")),
                show_col_types = FALSE) |>
    mutate(y = as.integer(.data[[m]]),
           world_region = fct_relevel(factor(world_region), "Africa"))

  # phylo sample = mapped entries only; duplicate tip col for the two effects
  d_phylo <- d |> filter(!is.na(tip_name)) |> mutate(phylo = tip_name)

  # per-marker tree prune + correlation matrix
  tips <- unique(d_phylo$tip_name)
  stopifnot(all(tips %in% tree$tip.label)) # guard 
  tree_pruned <- keep.tip(tree, tips)
  A <- vcv.phylo(tree_pruned, corr = TRUE)
  stopifnot(all(d_phylo$phylo %in% rownames(A)))

  f <- bf(as.formula(paste(
    m, "~ 1 + violent_external + year_scaled +",
    "(1 | world_region) + (1 | gr(phylo, cov = A)) + (1 | tip_name)"
  )))

  fit <- brm(f, data = d_phylo, data2 = list(A = A),
             family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 4000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99, max_treedepth = 12),
             seed = 52312, backend = "cmdstanr", silent = 2)

  fits_phylo[[m]] <- list(fit = fit, data = d_phylo, n = nrow(d_phylo),
                    n_tips = length(tips))
}
```

Per-marker convergence.

```{r diagnostics_phylo}

# per-marker convergence sweep — CATCH bad fits before trusting results
diag_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  tibble(
    Marker        = marker_labels[m],
    N             = fits_phylo[[m]]$n,
    n_tips        = fits_phylo[[m]]$n_tips,
    divergences   = sum(subset(nuts_params(fit), Parameter=="divergent__")$Value),
    max_rhat      = round(max(rhat(fit), na.rm=TRUE), 3),
    min_bulk_ess  = round(min(summary(fit)$fixed[,"Bulk_ESS"]), 0),
    min_tail_ess  = round(min(summary(fit)$fixed[,"Tail_ESS"]), 0)
  )
})
diag_tbl
write_csv(diag_tbl, here("data","model","external_noeHRAF","results","diagnostics_phylo.csv"))

```

Fixed effects table

```{r fixed_effects_phylo}

# one row per marker; each term formatted as "estimate [lo, hi]" — same shape as fixed_effects_baseline
fixed_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  fx  <- fixef(fit) # rows: Intercept, violent_external, year_scaled

  tibble(
    Marker           = marker_labels[m],
    N                = fits_phylo[[m]]$n,
    Intercept        = fmt(fx["Intercept","Estimate"], fx["Intercept","Q2.5"], fx["Intercept","Q97.5"]),
    violent_external = fmt(fx["violent_external","Estimate"], fx["violent_external","Q2.5"], fx["violent_external","Q97.5"]),
    year_scaled      = fmt(fx["year_scaled","Estimate"], fx["year_scaled","Q2.5"], fx["year_scaled","Q97.5"])
  )
})

fixed_tbl
write_csv(fixed_tbl, here("data","model","external_noeHRAF","results","fixed_effects_phylo.csv"))

```

Random effects table: one row per marker, sd(Intercept) for world_region, phylo (covariance-
structured tip effect), and tip_name (iid tip effect) — same "est [lo, hi]" format as baseline.

```{r random_effects_phylo}

sd_tbl <- map_dfr(markers_table_order, function(m) {
  fit       <- fits_phylo[[m]]$fit
  rf_region <- summary(fit)$random$world_region
  rf_phylo  <- summary(fit)$random$phylo
  rf_tip    <- summary(fit)$random$tip_name

  tibble(
    Marker          = marker_labels[m],
    N               = fits_phylo[[m]]$n,
    sd_world_region = fmt(rf_region["sd(Intercept)","Estimate"],
                           rf_region["sd(Intercept)","l-95% CI"],
                           rf_region["sd(Intercept)","u-95% CI"]),
    sd_phylo        = fmt(rf_phylo["sd(Intercept)","Estimate"],
                           rf_phylo["sd(Intercept)","l-95% CI"],
                           rf_phylo["sd(Intercept)","u-95% CI"]),
    sd_tip_name     = fmt(rf_tip["sd(Intercept)","Estimate"],
                           rf_tip["sd(Intercept)","l-95% CI"],
                           rf_tip["sd(Intercept)","u-95% CI"])
  )
})

sd_tbl
write_csv(sd_tbl, here("data","model","external_noeHRAF","results","random_effects_phylo.csv"))

```

Phylogenetic signal: proportion of tip-level variance that is phylogenetically structured
(sd_phylo, tree-respecting) vs. idiosyncratic per-tip (sd_tip_name). Computed per posterior
draw so the ratio itself carries uncertainty, not just from the two point estimates.

```{r phylo_signal}

phylo_signal_tbl <- map_dfr(markers_table_order, function(m) {
  fit   <- fits_phylo[[m]]$fit
  draws <- as_draws_df(fit, variable = c("sd_phylo__Intercept", "sd_tip_name__Intercept"))
  h2    <- draws$sd_phylo__Intercept^2 /
    (draws$sd_phylo__Intercept^2 + draws$sd_tip_name__Intercept^2)

  tibble(
    Marker       = marker_labels[m],
    N            = fits_phylo[[m]]$n,
    phylo_signal = fmt(mean(h2), quantile(h2, 0.025), quantile(h2, 0.975))
  )
})

phylo_signal_tbl
write_csv(phylo_signal_tbl, here("data","model","external_noeHRAF","results","phylo_signal.csv"))

```

AME + hypothesis table

```{r results_table}

results_tbl <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit
  d   <- fits_phylo[[m]]$data
  h <- hypothesis(fit, "violent_external > 0", alpha = 0.025)$hypothesis

  # AME: re_formula = NA excludes ALL random effects (region, phylo, tip)
  nd <- expand_grid(violent_external = c(0,1), year_scaled = 0,
                    world_region = unique(d$world_region))
  ame <- fit |> epred_draws(newdata = nd, re_formula = NA) |> ungroup() |> 
    group_by(.draw, violent_external) |>
    summarise(p = mean(.epred), .groups = "drop") |>
    pivot_wider(names_from = violent_external, values_from = p, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0) |> pull(ame)

  tibble(
    Marker = marker_labels[m], N = fits_phylo[[m]]$n,
    beta = h$Estimate, ci_lo = h$CI.Lower, ci_hi = h$CI.Upper,
    post_prob = h$Post.Prob, AME = mean(ame)
  )
}) |> mutate(across(where(is.numeric), \(x) round(x, 2)))

results_tbl
write_csv(results_tbl, here("data","model","external_noeHRAF","results","hypothesis_ame_phylo.csv"))

```

Plot AME distributions: baseline vs. phylo, to see how much the violent_external effect
shifts once phylogenetic + regional random effects are added.

```{r}

# AME draws for ALL markers under the phylo model (re_formula = NA, same as results_table above)
ame_all_phylo <- map_dfr(markers_table_order, function(m) {
  fit <- fits_phylo[[m]]$fit; d <- fits_phylo[[m]]$data
  nd <- expand_grid(violent_external = c(0,1), year_scaled = 0,
                    world_region = unique(d$world_region))
  fit |> epred_draws(newdata = nd, re_formula = NA) |> ungroup() |>
    group_by(.draw, violent_external) |>
    summarise(p = mean(.epred), .groups = "drop") |>
    pivot_wider(names_from = violent_external, values_from = p, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0, marker = m)
})

ame_compare <- bind_rows(
  ame_all_baseline |> mutate(model = "Baseline"),
  ame_all_phylo    |> mutate(model = "Phylogenetic")
) |> mutate(model = fct_relevel(model, "Baseline", "Phylogenetic"))

# marker order + labels from the baseline plot above, with N per model (fits/fits_phylo are
# separate objects now, so both are still in memory here — no need to read baseline N back from disk)
# parent markers get **bold** markdown around their label, rendered via
# ggtext::element_markdown() on the axis text below — same trick as p_base_rates.
n_labels_compare <- map_chr(markers, \(m) {
  lbl <- paste0(marker_labels[m], " (n=", fits[[m]]$n, "/", fits_phylo[[m]]$n, ")")
  if (marker_level[m] == "parent") paste0("**", lbl, "**") else lbl
}) |> setNames(markers)

p_ame_compare <- ame_compare |>
  mutate(marker_label = factor(n_labels_compare[marker], levels = n_labels_compare[marker_order])) |>
  ggplot(aes(x = ame, y = marker_label, fill = model)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               position = position_dodge(width = 0.6), scale = 1.1,
               color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c(Baseline = "#C8839E", Phylogenetic = "#6B9AC4")) +
  labs(x = "Average marginal effect (probability)", y = NULL, fill = NULL) +
  theme_bw(base_size = 16) +
  theme(
    plot.title    = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16),
    axis.title.x  = element_text(size = 17),
    axis.text.y   = element_markdown(size = 15),
    axis.text.x   = element_text(size = 14),
    legend.position = "top",
    plot.margin   = margin(12, 16, 12, 16)
  )

ggsave(here("data","model","external_noeHRAF","results","external_noeHRAF_ame_baseline_vs_phylo.pdf"),
       p_ame_compare, width = 9, height = 7, device = "pdf")
p_ame_compare

```

## DEEP DIVE: TATTOOS/SCARIFICATION ##

Per-tip breakdown: raw (unpooled) rate vs. model-estimated (partially pooled) rate, grouped by
language family (first dot-separated segment of the tip name) — same spirit as
region_effects_baseline. Model estimate combines Intercept + phylo + tip_name via epred_draws,
so it's properly anchored (unlike the raw ranef offset, whose sign/level is confounded with the
Intercept for a phylogenetically-correlated random effect — see discussion above).

```{r tip_effects_tattoos_scarification}

m   <- "tattoos_scarification"
fit <- fits_phylo[[m]]$fit
d   <- fits_phylo[[m]]$data

# rebuild A for this marker's tips specifically (self-contained; don't rely on the loop-leaked
# `A` from fit_phylo above, which only happens to be correct because this marker is last in `markers`)
tree_pruned_m <- keep.tip(tree, unique(d$tip_name))
A <- vcv.phylo(tree_pruned_m, corr = TRUE)

raw <- d |> group_by(tip_name) |>
  summarise(n = n(), raw_rate = mean(y), .groups = "drop")

nd <- tibble(tip_name = unique(d$tip_name), phylo = unique(d$tip_name),
             violent_external = 0, year_scaled = 0, world_region = d$world_region[1])

model_est <- fit |> epred_draws(newdata = nd, re_formula = ~ (1 | gr(phylo, cov = A)) + (1 | tip_name)) |>
  ungroup() |>
  group_by(tip_name) |>
  summarise(model_estimate = mean(.epred),
            ci_lo = quantile(.epred, 0.025),
            ci_hi = quantile(.epred, 0.975), .groups = "drop")

tip_tbl <- raw |>
  left_join(model_est, by = "tip_name") |>
  #filter(n > 5) |>
  mutate(family = str_split_i(tip_name, "\\.", 1), .before = 1) |>
  select(family, tip_name, n, raw_rate, model_estimate, ci_lo, ci_hi) |>
  arrange(family, desc(model_estimate)) |>
  mutate(across(c(raw_rate, model_estimate, ci_lo, ci_hi), \(x) round(x, 2)))

tip_tbl

```

Aggregated to language family — the level we actually report at. One row per family: number
of tips, total observations, and mean raw/model-estimated rate (min/max kept to show spread).

```{r family_effects_tattoos_scarification}

family_tbl <- tip_tbl |>
  group_by(family) |>
  summarise(
    total_n       = sum(n),
    n_tips        = n(),
    mean_raw_rate = round(sum(raw_rate * n) / sum(n), 3),
    mean_estimate = round(mean(model_estimate), 3),
    #min_estimate  = round(min(model_estimate), 2),
    #max_estimate  = round(max(model_estimate), 2),
    .groups = "drop"
  ) |>
  arrange(desc(total_n))

family_tbl
write_csv(family_tbl, here("data","model","external_noeHRAF","results","tip_effects_tattoos_scarification.csv"))

```

Tree visualization (not finished.)

```{r tree_annotated_tattoos_scarification}

m <- "tattoos_scarification"
d <- fits_phylo[[m]]$data

tip_data <- d |> group_by(tip_name) |>
  summarise(n = n(),
            tattoo_rate  = mean(y),
            violent_rate = mean(violent_external),
            .groups = "drop")

tree_m <- keep.tip(tree, unique(d$tip_name))
n_tip  <- Ntip(tree_m)

pdf(here("data","model","external_noeHRAF","results","tree_tattoos_scarification.pdf"),
    width = 10, height = n_tip * 0.15 + 2)

plot.phylo(tree_m, use.edge.length = FALSE, show.tip.label = TRUE,
           cex = 0.4, label.offset = 3, no.margin = FALSE)

lastPP <- get("last_plot.phylo", envir = .PlotPhyloEnv)
tip_y  <- lastPP$yy[1:n_tip]
tip_x  <- max(lastPP$xx[1:n_tip]) + 1.5  # start of annotation columns, past tip labels

tip_ord <- data.frame(tip_name = tree_m$tip.label, y = tip_y) |>
  left_join(tip_data, by = "tip_name")

col_tattoo  <- gray(1 - tip_ord$tattoo_rate)
col_violent <- gray(1 - tip_ord$violent_rate)

points(rep(tip_x,     n_tip), tip_ord$y, pch = 15, cex = 1, col = col_tattoo)
points(rep(tip_x + 1, n_tip), tip_ord$y, pch = 15, cex = 1, col = col_violent)

text(tip_x,     n_tip + 2, "Tattoos", cex = 0.6, srt = 90, adj = 0)
text(tip_x + 1, n_tip + 2, "Violent ext.", cex = 0.6, srt = 90, adj = 0)

dev.off()

```
