---
title: "Phylogenetic model: extra ritual group markers ~ violent external conflict"
output:
  html_document:
    toc: true
    toc_float: true
---

This script fits a Bayesian phylogenetic logistic regression using the ASJP world
language tree as a covariance structure. Rather than aggregating DRH entries to one
value per language tip, we keep all 240 entry-level observations and let the
phylogenetic random effect handle the non-independence between entries that share a
language.

**Model:** `extra_ritual_group_markers ~ violent_external + (1 | gr(tip_name, cov = A))`

The `gr(tip_name, cov = A)` term is a Gaussian process over the tree: languages that
share more recent common ancestry have more correlated random intercepts. This plays
the same role as world_region in the existing models, but uses linguistic genealogy
rather than geography.

```{r setup, message=FALSE, warning=FALSE}
library(pacman)
p_load(brms, ape, tidyverse, here, HDInterval)

set.seed(1342)
```

## 1. Load data and tree

```{r load}
data <- read_csv(here("data", "phylo_input", "extra_ritual_group_markers.csv"))
tree <- read.tree(here("data", "jaeger2018", "world.tre"))

cat("Entries:", nrow(data), "\n")
cat("Unique tips in data:", n_distinct(data$tip_name), "\n")
cat("Tips in full tree:  ", length(tree$tip.label), "\n")
```

## 2. Prune tree to the tips we have data for

BayesTraits and brms both require that the covariance matrix covers exactly the tips
present in the data — no more, no less.

```{r prune}
tips_in_data <- unique(data$tip_name)
tree_pruned  <- keep.tip(tree, tips_in_data)

cat("Tips in pruned tree:", length(tree_pruned$tip.label), "\n")
stopifnot(setequal(tree_pruned$tip.label, tips_in_data))
```

## 3. Phylogenetic covariance matrix

`vcv.phylo(..., corr = TRUE)` gives a correlation matrix (diagonal = 1) where
off-diagonal entries are the proportion of total depth shared between two tips — i.e.,
how recently they diverged. brms uses this as the prior covariance for the
language-level random intercepts.

```{r cov_matrix}
A <- vcv.phylo(tree_pruned, corr = TRUE)
dim(A)  # should be 147 x 147
```

## 4. Baseline model (no phylogeny)

Fitting the non-phylogenetic model first gives us a reference for what the raw
association between violent conflict and ritual markers looks like, ignoring linguistic
relatedness entirely.

```{r fit_baseline, cache=TRUE, message=FALSE, warning=FALSE}
fit_base <- brm(
    extra_ritual_group_markers ~ 1 + violent_external,
    data   = data,
    family = bernoulli(link = logit),
    prior  = c(
        prior(normal(0, 1.5), class = Intercept),
        prior(normal(0, 1),   class = b)
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4,
    control = list(adapt_delta = .999, max_treedepth = 20),
    seed   = 1342,
    silent = 2
)
```

## 5. Phylogenetic model

The `gr(tip_name, cov = A)` term adds a language-level random intercept whose prior
covariance is the phylogenetic correlation matrix `A`. The `sd` parameter for this term
captures how much variance is attributable to shared linguistic ancestry — analogous to
Pagel's λ.

Note: entries that share the same `tip_name` (same language) get the same phylogenetic
random intercept, but still have independent residual variation. This is what lets us
use all 314 observations rather than collapsing to 147.

```{r fit_phylo, cache=TRUE, message=FALSE, warning=FALSE}
fit_phylo <- brm(
    extra_ritual_group_markers ~ 1 + violent_external + (1 | gr(tip_name, cov = A)),
    data   = data,
    data2  = list(A = A),
    family = bernoulli(link = logit),
    prior  = c(
        prior(normal(0, 1.5), class = Intercept),
        prior(normal(0, 1),   class = b),
        prior(exponential(1), class = sd)
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4,
    control = list(adapt_delta = .999, max_treedepth = 20),
    seed   = 1342,
    silent = 2
)
```

## 6. Results

```{r results}
summary(fit_base) # violent_external: 1.05 (0.61, 1.53)
summary(fit_phylo) # violent_external: 1.07 (0.59, 1.56) <- basically the same, slightly more uncertainty 
```

### Does the violent_external effect survive phylogenetic control?

```{r hypothesis}
# one-sided: does violent conflict positively predict ritual markers?
hypothesis(fit_base,  "violent_external > 0") # 1.05 (0.67, 1.44)
hypothesis(fit_phylo, "violent_external > 0") # 1.07 (0.67, 1.48)
```

### How strong is the phylogenetic signal?

The `sd(tip_name)` parameter from the phylogenetic model tells you how much
between-language variance is explained by the tree. If it's large relative to the
overall variance, language ancestry is doing a lot of work. Here it is not super large.

```{r phylo_signal}
# Extract sd of the phylogenetic random effect
post <- as_draws_df(fit_phylo)
sd_phylo <- post$`sd_tip_name__Intercept`

cat("Phylogenetic sd — median:", round(median(sd_phylo), 3),
    "  95% HDI:", round(hdi(sd_phylo, credMass = 0.95), 3), "\n")

```

### Compare violent_external estimate: baseline vs. phylogenetic
- Relatively minor effect of adding the phylogenetic control.
- This seems surprising, need to verify that the covariances make sense and are included correctly. 

```{r compare}
b_base  <- as_draws_df(fit_base)$b_violent_external
b_phylo <- as_draws_df(fit_phylo)$b_violent_external

tibble(
    model  = c("baseline", "phylogenetic"),
    median = c(median(b_base), median(b_phylo)),
    lo95   = c(hdi(b_base)[1], hdi(b_phylo)[1]),
    hi95   = c(hdi(b_base)[2], hdi(b_phylo)[2])
) |> mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## 7. Control for year as a covariate:
- This does almost nothing to our results. 
- Year is a really small effect.

```{r fit_phylo_year, cache=TRUE}
fit_phylo_year <- brm(
    extra_ritual_group_markers ~ 1 + violent_external + year_scaled + world_region + 
        (1 | gr(tip_name, cov = A)),
    data   = data,
    data2  = list(A = A),
    family = bernoulli(link = logit),
    prior  = c(
        prior(normal(0, 1.5), class = Intercept),
        prior(normal(0, 1),   class = b),
        prior(exponential(1), class = sd)
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4,
    control = list(adapt_delta = .999, max_treedepth = 20),
    seed   = 1342
)
summary(fit_phylo_year)
hypothesis(fit_phylo_year, "violent_external > 0")
```
