---
title: "5_societal_scale"
output: html_document
---


```{r}
library(pacman)
p_load(brms, tidyverse, here)
set.seed(12532)

d <- read_csv(here("data","preprocessed","answerset_large.csv"), show_col_types = FALSE)
d %>% glimpse()

# fit1/fit2 (state only) are compared with fit3/fit4 (state + conflict), so they are
# fitted on the same rows: entries with conflict coded
d_conflict <- d |> filter(!is.na(violent_external))

dir.create(here("data","model","societal_scale","results"), recursive = TRUE, showWarnings = FALSE)
```

```{r}

priors <- c(
  prior(normal(0, 1.5), class = Intercept), 
  prior(normal(0, 1),   class = b),
  prior(exponential(1), class = sd) 
)

```


Model 1: marker ~ state
Extra-ritual in-group markers (same entries as model 3)

```{r}

fit1 <- brm(extra_ritual_group_markers ~ 1 + state + year_scaled + (1|world_region), 
            data = d_conflict, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

summary(fit1)

```

Model 2: marker ~ state
Permanent scarring (same entries as model 4)

```{r}

fit2 <- brm(permanent_scarring ~ 1 + state + year_scaled + (1|world_region), 
            data = d_conflict, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

summary(fit2)

```

Model 3: marker ~ state + violent_external
Extra-ritual in-group markers

```{r}

fit3 <- brm(extra_ritual_group_markers ~ 1 + state + violent_external + year_scaled + (1|world_region), 
            data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

summary(fit3)

```

Model 4: marker ~ state + violent_external
Permanent scarring

```{r}

fit4 <- brm(permanent_scarring ~ 1 + state + violent_external + year_scaled + (1|world_region), 
            data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

summary(fit4)

```

Model 5: violent_external ~ state 

```{r}

fit5 <- brm(violent_external ~ 1 + state + year_scaled + (1|world_region), 
            data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

summary(fit5)

```

Reporting order, shared by the diagnostics and fixed-effects tables below: the conflict
model first, then each marker's total (state only) and conditional (state + conflict) model
side by side.

```{r state_fits}

state_fits <- list(
  "Violent external conflict ~ state" = fit5,
  "Extra-ritual in-group markers ~ state" = fit1,
  "Extra-ritual in-group markers ~ state + conflict" = fit3,
  "Permanent scarring ~ state" = fit2,
  "Permanent scarring ~ state + conflict" = fit4
)

```

Convergence

```{r diagnostics}

diag_tbl <- map_dfr(names(state_fits), function(nm) {
  fit <- state_fits[[nm]]
  fixed <- summary(fit)$fixed
  sds <- bind_rows(map(summary(fit)$random, as.data.frame))

  tibble(
    Model = nm,
    N = nobs(fit),
    divergences = sum(subset(nuts_params(fit), Parameter=="divergent__")$Value),
    max_rhat = round(max(rhat(fit), na.rm=TRUE), 3),
    min_bulk_ess = round(min(c(fixed[,"Bulk_ESS"], sds[,"Bulk_ESS"])), 0),
    min_tail_ess = round(min(c(fixed[,"Tail_ESS"], sds[,"Tail_ESS"])), 0)
  )
})
diag_tbl
write_csv(diag_tbl, here("data","model","societal_scale","results","diagnostics.csv"))

```

Posterior predictive check: observed prevalence against its posterior predictive
interval, split by statehood (the predictor shared by all five models; external conflict
is the outcome of the first one).

```{r ppc}

ppc <- map_dfr(names(state_fits), function(nm) {
  fit <- state_fits[[nm]]
  d <- fit$data
  y <- d[[1]] # brms puts the response first in the model frame
  yrep <- posterior_predict(fit, ndraws = 1000)
  state <- ifelse(d$state == 1, "Yes", "No")

  map_dfr(c("No", "Yes"), function(g) {
    idx <- which(state == g)
    prop <- rowMeans(yrep[, idx, drop = FALSE])
    tibble(
      Model = nm,
      state = g,
      observed = mean(y[idx]),
      lo = as.numeric(quantile(prop, 0.025)),
      hi = as.numeric(quantile(prop, 0.975))
    )
  })
}) |>
  mutate(Model = factor(Model, levels = rev(names(state_fits))))

ggplot(ppc, aes(y = Model, colour = state)) +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 3, alpha = 0.4,
                 position = position_dodge(width = 0.5)) +
  geom_point(aes(x = observed), size = 2,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c(No = "#8FBF9F", Yes = "#D98B72")) +
  labs(x = "Outcome prevalence", y = NULL, colour = "State",
       subtitle = "Bars: 95% posterior predictive interval. Points: observed proportion.") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

```

Fixed effects. A term absent from a model is written as NA here and rendered as a dash in
the LaTeX table; external violent conflict is the outcome of the first model, not a term.

```{r fixed_effects}

fmt <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)

fmt_term <- function(fx, term) {
  if (!term %in% rownames(fx)) return(NA_character_)
  fmt(fx[term,"Estimate"], fx[term,"Q2.5"], fx[term,"Q97.5"])
}

fixed_tbl <- map_dfr(names(state_fits), function(nm) {
  fit <- state_fits[[nm]]
  fx <- fixef(fit)

  tibble(
    Model = nm,
    N = nobs(fit),
    Intercept = fmt_term(fx, "Intercept"),
    state = fmt_term(fx, "state"),
    violent_external = fmt_term(fx, "violent_external"),
    year_scaled = fmt_term(fx, "year_scaled")
  )
})

fixed_tbl
write_csv(fixed_tbl, here("data","model","societal_scale","results","state_fixed_effects.csv"))

```

Directional hypothesis tests, same form as the hypothesis tables in analyses 1-4.

```{r hypothesis_table}

hyp_tbl <- map_dfr(names(state_fits), function(nm) {
  fit <- state_fits[[nm]]
  terms <- intersect(c("state", "violent_external"), rownames(fixef(fit)))

  map_dfr(terms, function(tm) {
    # directions follow Section 3.4: statehood lowers rates, conflict raises them
    dir <- if (tm == "state") "< 0" else "> 0"
    h <- hypothesis(fit, paste(tm, dir), alpha = 0.025)$hypothesis

    tibble(
      Model = nm,
      hypothesis = paste(tm, dir),
      beta = round(h$Estimate, 2),
      ci_lo = round(h$CI.Lower, 2),
      ci_hi = round(h$CI.Upper, 2),
      post_prob = round(h$Post.Prob, 2)
    )
  })
})

hyp_tbl
write_csv(hyp_tbl, here("data","model","societal_scale","results","state_hypothesis.csv"))

```
