---
title: "6_conditional"
output: html_document
---

Analysis of marker ~ external violent conflict, restricted to entries where the parent
question ("Are extra-ritual in-group markers present") is answered "Yes".

The six markers modeled here are child questions of that parent, and a coder only sees
them when the parent is "Yes". 1_curate_data.py therefore infers "No" for every child of
a "No" parent, which makes ~62% of each child marker's rows in the main analysis a copy
of the parent's answer rather than an observed coding. Dropping those rows (done in
4_curate_models.py) changes the question from "does conflict predict that markers are
present at all?" to "given that markers are present, does conflict predict which kind?".

Note that the parent is itself an outcome of conflict, so this is a different estimand
rather than a corrected version of 1_external.Rmd — conditioning on it is conditioning on
a post-treatment variable. The two analyses are reported side by side, not one in place
of the other.

Baseline model only. The phylogenetic variant is deferred: this sample has 164 entries
across 83 language tips (mostly singletons), so the tip-level random intercept would
absorb most of the conflict effect and the result would be near-uninformative.

```{r}
library(pacman)
p_load(brms, tidyverse, here, tidybayes, ggdist)
set.seed(12532)

dir.create(here("data","model","external_markers_present","results"),
           recursive = TRUE, showWarnings = FALSE)
```

```{r}

# the six sub-questions of "Are extra-ritual in-group markers present", ordered permanent
# first then transitory (the contrast of interest) rather than alphabetically. Unlike
# 1_external.Rmd there is no parent/sub distinction to encode here — every marker in this
# analysis is a child question — so the marker_level machinery from that file is dropped.
markers <- c(
  "circumcision", "tattoos_scarification",       # permanent
  "dress", "food_taboos", "hair", "ornaments"    # transitory
)

marker_labels <- c(
  circumcision="Circumcision", tattoos_scarification="Tattoos/Scarification",
  dress="Dress", food_taboos="Food Taboos", hair="Hair", ornaments="Ornaments"
)

marker_type <- c(
  circumcision          = "Permanent",
  tattoos_scarification = "Permanent",
  dress                 = "Transitory",
  food_taboos           = "Transitory",
  hair                  = "Transitory",
  ornaments             = "Transitory"
)

# identical to 1_external.Rmd so the two analyses stay comparable
priors <- c(
  prior(normal(0, 1.5), class = Intercept),
  prior(normal(0, 1),   class = b),
  prior(exponential(1), class = sd)
)

fits <- list()
for (m in markers) {
  d <- read_csv(here("data","model","external_markers_present","input", paste0(m, ".csv")),
                show_col_types = FALSE) |>
    mutate(world_region = fct_relevel(factor(world_region), "Africa"))

  f <- bf(as.formula(paste(m, "~ 1 + violent_external + year_scaled + (1 | world_region)")))

  fit <- brm(f, data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 4000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

  fits[[m]] <- list(fit = fit, data = d, n = nrow(d))
}

```

Convergence

```{r diagnostics_baseline}

diag_tbl <- map_dfr(markers, function(m) {
  fit <- fits[[m]]$fit
  tibble(
    Marker       = marker_labels[m],
    N            = fits[[m]]$n,
    divergences  = sum(subset(nuts_params(fit), Parameter=="divergent__")$Value),
    max_rhat     = round(max(rhat(fit), na.rm=TRUE), 3),
    min_bulk_ess = round(min(summary(fit)$fixed[,"Bulk_ESS"]), 0),
    min_tail_ess = round(min(summary(fit)$fixed[,"Tail_ESS"]), 0)
  )
})
diag_tbl
write_csv(diag_tbl, here("data","model","external_markers_present","results","diagnostics_baseline.csv"))

```

Fixed effects table

```{r fixed_effects_baseline}

fmt <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)

fixed_tbl <- map_dfr(markers, function(m) {
  fit <- fits[[m]]$fit
  fx  <- fixef(fit) # rows: Intercept, violent_external, year_scaled

  tibble(
    Marker           = marker_labels[m],
    N                = fits[[m]]$n,
    Intercept        = fmt(fx["Intercept","Estimate"], fx["Intercept","Q2.5"], fx["Intercept","Q97.5"]),
    violent_external = fmt(fx["violent_external","Estimate"], fx["violent_external","Q2.5"], fx["violent_external","Q97.5"]),
    year_scaled      = fmt(fx["year_scaled","Estimate"], fx["year_scaled","Q2.5"], fx["year_scaled","Q97.5"])
  )
})

fixed_tbl
write_csv(fixed_tbl, here("data","model","external_markers_present","results","fixed_effects_baseline.csv"))

```

Random effects (world region). Per-region breakdown is deliberately not reported for this
analysis: with 164 entries spread over 10 regions (5 in Central Eurasia, 5 in
Oceania-Australia) the raw regional rates are too thin to interpret.

```{r random_effects_baseline}

sd_tbl <- map_dfr(markers, function(m) {
  fit <- fits[[m]]$fit
  rf  <- summary(fit)$random$world_region  # row: sd(Intercept)

  tibble(
    Marker          = marker_labels[m],
    N               = fits[[m]]$n,
    sd_world_region = fmt(rf["sd(Intercept)","Estimate"],
                          rf["sd(Intercept)","l-95% CI"],
                          rf["sd(Intercept)","u-95% CI"])
  )
})

sd_tbl
write_csv(sd_tbl, here("data","model","external_markers_present","results","random_effects_baseline.csv"))

```

Hypothesis test + AME. Same construction as 1_external.Rmd so the two are directly
comparable.

```{r hypothesis_ame_baseline}

results_tbl <- map_dfr(markers, function(m) {
  fit <- fits[[m]]$fit
  d   <- fits[[m]]$data

  h <- hypothesis(fit, "violent_external > 0", alpha = 0.025)$hypothesis

  nd <- expand_grid(violent_external = c(0,1), year_scaled = 0,
                    world_region = unique(d$world_region))
  ame <- fit |> epred_draws(newdata = nd, re_formula = NA) |> ungroup() |>
    group_by(.draw, violent_external) |>
    summarise(p = mean(.epred), .groups = "drop") |>
    pivot_wider(names_from = violent_external, values_from = p, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0) |> pull(ame)

  tibble(
    Marker = marker_labels[m],
    Type   = marker_type[m],
    N      = fits[[m]]$n,
    beta   = h$Estimate,
    ci_lo  = h$CI.Lower,
    ci_hi  = h$CI.Upper,
    post_prob = h$Post.Prob,
    AME    = mean(ame)
  )
}) |> mutate(across(where(is.numeric), \(x) round(x, 2)))

results_tbl
write_csv(results_tbl, here("data","model","external_markers_present","results","hypothesis_ame_baseline.csv"))

```

AME draws for all markers, kept as draws so both plots below can be built from them (the
base rates are the cond0/cond1 pair before collapsing to their difference). Same
construction as the AME column of the table above.

```{r ame_draws}

ame_all <- map_dfr(markers, function(m) {
  fit <- fits[[m]]$fit; d <- fits[[m]]$data
  nd <- expand_grid(violent_external = c(0,1), year_scaled = 0,
                    world_region = unique(d$world_region))
  fit |> epred_draws(newdata = nd, re_formula = NA) |> ungroup() |>
    group_by(.draw, violent_external) |>
    summarise(p = mean(.epred), .groups = "drop") |>
    pivot_wider(names_from = violent_external, values_from = p, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0, marker = m)
})

# two orderings, following 1_external.Rmd: the base-rates plot sorts by prevalence under
# the focal condition, the AME plot by effect size. The permanent/transitory split is
# deliberately not encoded in the labels — bold already means "top-level question" in the
# figures of 1_external.Rmd, and the split is carried by the Type column of results_tbl.
marker_order <- ame_all |> group_by(marker) |>
  summarise(md = median(ame)) |> arrange(md) |> pull(marker)

marker_order_prevalence <- ame_all |> group_by(marker) |>
  summarise(md = median(cond1)) |> arrange(md) |> pull(marker)

n_labels <- map_chr(markers, \(m) {
  paste0(marker_labels[m], " (n=", fits[[m]]$n, ")")
}) |> setNames(markers)

```

Estimated base rates: predicted probability under violent_external = 0 vs. = 1. Note that
these are conditional prevalences (among groups that have markers at all), so they sit
much higher than the corresponding plot in 1_external.Rmd and are not comparable to it.

```{r base_rates_baseline}

p_base_rates <- ame_all |>
  pivot_longer(c(cond0, cond1), names_to = "condition", values_to = "p") |>
  mutate(condition = fct_recode(condition,
                                "No violent conflict" = "cond0",
                                "Violent conflict"    = "cond1"),
         marker_label = factor(n_labels[marker], levels = n_labels[marker_order_prevalence])) |>
  ggplot(aes(x = p, y = marker_label, fill = condition)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               position = position_dodge(width = 0.3), scale = 1.0,
               color = "black", alpha = 0.7) +
  # headroom at the top so the last marker's density is not clipped by the panel
  scale_y_discrete(expand = expansion(add = c(0.6, 1.0))) +
  scale_fill_manual(values = c("No violent conflict" = "#8FBF9F",
                               "Violent conflict"    = "#D98B72")) +
  labs(x = "Estimated prevalence (given markers present)", y = NULL, fill = NULL) +
  theme_bw(base_size = 16) +
  theme(
    axis.title.x    = element_text(size = 17),
    axis.text.y     = element_text(size = 15),
    axis.text.x     = element_text(size = 14),
    legend.position = "top",
    plot.margin     = margin(12, 16, 12, 16)
  )

ggsave(here("data","model","external_markers_present","results","markers_present_base_rates_baseline.pdf"),
       p_base_rates, width = 9, height = 7, device = "pdf")
p_base_rates

```

Average marginal effect of external violent conflict per marker.

```{r ame_baseline}

p_ame <- ame_all |>
  mutate(marker_label = factor(n_labels[marker], levels = n_labels[marker_order])) |>
  ggplot(aes(x = ame, y = marker_label)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi, scale = 0.9,
               fill = "#6B9AC4", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  # headroom at the top so the last marker's density is not clipped by the panel
  scale_y_discrete(expand = expansion(add = c(0.6, 1.0))) +
  labs(x = "Average marginal effect (probability)", y = NULL) +
  theme_bw(base_size = 16) +
  theme(
    axis.title.x    = element_text(size = 17),
    axis.text.y     = element_text(size = 15),
    axis.text.x     = element_text(size = 14),
    plot.margin     = margin(12, 16, 12, 16)
  )

ggsave(here("data","model","external_markers_present","results","markers_present_ame_baseline.pdf"),
       p_ame, width = 9, height = 7, device = "pdf")
p_ame

```
