---
title: "Phylogenetic model: extra ritual group markers ~ violent external conflict"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(pacman)
p_load(brms, ape, tidyverse, here, HDInterval)

set.seed(1342)
```

## 1. Load data and tree

```{r load}
data <- read_csv(here("data", "phylo_input", "tattoos_scarification.csv"))
tree <- read.tree(here("asjp", "raw", "world.tre"))

cat("Entries:", nrow(data), "\n")
cat("Unique tips in data:", n_distinct(data$tip_name), "\n")
cat("Tips in full tree:  ", length(tree$tip.label), "\n")
```
## 2. Prune tree to the tips we have data for

BayesTraits and brms both require that the covariance matrix covers exactly the tips
present in the data — no more, no less.

```{r prune}
tips_in_data <- unique(data$tip_name)
tree_pruned  <- keep.tip(tree, tips_in_data)

cat("Tips in pruned tree:", length(tree_pruned$tip.label), "\n")
stopifnot(setequal(tree_pruned$tip.label, tips_in_data))

```

## 3. Phylogenetic covariance matrix

`vcv.phylo(..., corr = TRUE)` gives a correlation matrix (diagonal = 1) where
off-diagonal entries are the proportion of total depth shared between two tips — i.e.,
how recently they diverged. brms uses this as the prior covariance for the
language-level random intercepts.

```{r cov_matrix}
A <- vcv.phylo(tree_pruned, corr = TRUE)
dim(A)  # should be 147 x 147
```

## 4. Baseline model (no phylogeny)

```{r fit_baseline, cache=TRUE, message=FALSE, warning=FALSE}
fit_base <- brm(
    tattoos_scarification ~ 1 + violent_external,
    data   = data,
    family = bernoulli(link = logit),
    prior  = c(
        prior(normal(0, 1.5), class = Intercept),
        prior(normal(0, 1),   class = b)
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4,
    control = list(adapt_delta = .95, max_treedepth = 15),
    seed   = 1342,
    silent = 2
)
```

## 5. Phylogenetic model

```{r fit_phylo, cache=TRUE, message=FALSE, warning=FALSE}

fit_phylo <- brm(
    tattoos_scarification ~ 1 + violent_external + (1 | gr(tip_name, cov = A)),
    data   = data,
    data2  = list(A = A),
    family = bernoulli(link = logit),
    prior  = c(
        prior(normal(0, 1.5), class = Intercept),
        prior(normal(0, 1),   class = b),
        prior(exponential(1), class = sd)
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4,
    control = list(adapt_delta = .999, max_treedepth = 20),
    seed   = 1342,
    silent = 2
)

```

## 6. Results

```{r results}
#summary(fit_base) # violent_external: 0.85 (0.10, 1.59)
summary(fit_phylo) # violent_external: 0.85 (0.10, 1.60) <- basically the same.
```

```{r}

plot(fit_phylo, N = 2, ask = FALSE)

```

### Does the violent_external effect survive phylogenetic control?

```{r hypothesis}
# one-sided: does violent conflict positively predict ritual markers?
hyp_base <- hypothesis(fit_base,  "violent_external > 0") # 0.85 (0.24, 1.46)
hyp_phylo <- hypothesis(fit_phylo, "violent_external > 0") # 0.85 (0.22, 1.48)
```

```{r}

plot(hyp_phylo)

```

# Phylogenetic variation (13%); although not an expert here.

The so called phylogenetic signal (often symbolize by λ) can be computed with the hypothesis method and is ~.13 here.

For Bernoulli models there is no sigma — but the latent logistic distribution underlying the Bernoulli does have an implicit residual variance. The standard approach is to use the latent variable formulation, where the logistic distribution has a fixed residual variance of π²/3 ≈ 3.29.

This is analogous to Pagel's λ but on the latent logit scale — it tells you what proportion of the latent variance is attributable to phylogenetic structure. It's sometimes called the phylogenetic heritability in this context.


```{r}
hyp <- "sd_tip_name__Intercept^2 / (sd_tip_name__Intercept^2 + 3.289868) = 0"
(hyp <- hypothesis(fit_phylo, hyp, class = NULL))
plot(hyp)
```

### Compare violent_external estimate: baseline vs. phylogenetic
- Relatively minor effect of adding the phylogenetic control.
- This seems surprising, need to verify that the covariances make sense and are included correctly. 

```{r compare}
b_base  <- as_draws_df(fit_base)$b_violent_external
b_phylo <- as_draws_df(fit_phylo)$b_violent_external

tibble(
    model  = c("baseline", "phylogenetic"),
    median = c(median(b_base), median(b_phylo)),
    lo95   = c(hdi(b_base)[1], hdi(b_phylo)[1]),
    hi95   = c(hdi(b_base)[2], hdi(b_phylo)[2])
) |> mutate(across(where(is.numeric), \(x) round(x, 3)))
```

# Adding repeated measurements

We probably do not have enough observations per tip (most are unique) to properly estimate this model, but trying just as a robusness/sanity check.

```{r}

## 5b. Phylogenetic model with repeated measurements
data <- data %>% mutate(tip_name_rep = tip_name)
fit_phylo_repeat <- brm(
    tattoos_scarification ~ 1 + violent_external + 
        (1 | gr(tip_name, cov = A)) + 
        (1 | tip_name_rep),
    data   = data,
    data2  = list(A = A),
    family = bernoulli(link = logit),
    prior  = c(
        prior(normal(0, 1.5), class = Intercept),
        prior(normal(0, 1),   class = b),
        prior(exponential(1), class = sd)
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4,
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    seed   = 1342,
    silent = 2
)

```

```{r}
hyp <- paste(
    "sd_tip_name__Intercept^2 /",
    "(sd_tip_name__Intercept^2 + sd_tip_name_rep__Intercept^2 + 3.289868) = 0"
)
hypothesis(fit_phylo_repeat, hyp, class = NULL)
```
```{r}

summary(fit_phylo_repeat)

```

Basically does not change much, probably better to stick with the other model we have anyways then.

## 7. Control for year + region as well as covariates. 
- This does almost nothing to our results. 
- Year is a really small effect.

```{r fit_phylo_year, cache=TRUE}
fit_phylo_year <- brm(
    tattoos_scarification ~ 1 + violent_external + year_scaled + world_region + 
        (1 | gr(tip_name, cov = A)),
    data   = data,
    data2  = list(A = A),
    family = bernoulli(link = logit),
    prior  = c(
        prior(normal(0, 1.5), class = Intercept),
        prior(normal(0, 1),   class = b),
        prior(exponential(1), class = sd)
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4,
    control = list(adapt_delta = .999, max_treedepth = 20),
    seed   = 1342
)

```

```{r}
summary(fit_phylo_year)
```

```{r}
hypothesis(fit_phylo_year, "violent_external > 0")
```

