---
title: "7_combined"
output: html_document
---

A single "shared" model over the six child markers of "Are extra-ritual in-group markers
present", with marker type (permanent vs. transitory) as a moderator of the external
conflict effect. Where 1_external.Rmd and 6_conditional.Rmd fit one model per marker and
leave the reader to compare eight overlapping intervals, this estimates the
permanent-minus-transitory difference directly, as a single posterior quantity.

The same six markers and the same model are fitted to two samples:

  Full sample      all entries with a coding on the marker (data/model/external/input)
  Markers present  only entries whose parent question is "Yes", i.e. dropping the ~62%
                   of rows whose child answer was inherited from the parent rather than
                   observed (data/model/external_markers_present/input)

Holding the marker set fixed across the two means the only thing that differs is the
inherited rows, so the change in the contrast measures how much the coding hierarchy
inflates it.

Two limitations to keep in view when reading the output:

  - Only two of the six markers are permanent (circumcision, tattoos/scarification), so
    the model cannot fully separate "permanent markers behave differently" from "these
    two markers behave differently". The by-marker random slope propagates that into the
    contrast interval, which is why a wide interval is the expected result rather than a
    failure of the fit.
  - Restricting to parent = "Yes" conditions on a variable that is itself an outcome of
    conflict, so the markers-present fit is a different estimand (compositional) rather
    than a corrected version of the full-sample fit. Neither supersedes the other.

The long frame is built here rather than in 4_curate_models.py, deliberately: it is a pure
reshape of per-marker CSVs that already exist, and writing it to disk would create a second
copy of the same rows that could drift out of sync with them.

This file is exploratory and deliberately writes nothing to disk — no results CSVs, no
figures. It is not (yet) part of the paper; 6_conditional.Rmd is. Everything below is
printed inline only. If this analysis is adopted, add the saves then.

```{r}
library(pacman)
p_load(brms, tidyverse, here, tidybayes, ggdist)
set.seed(12532)
```

```{r}

markers <- c(
  "circumcision", "tattoos_scarification",       # permanent
  "dress", "food_taboos", "hair", "ornaments"    # transitory
)

marker_labels <- c(
  circumcision="Circumcision", tattoos_scarification="Tattoos/Scarification",
  dress="Dress", food_taboos="Food Taboos", hair="Hair", ornaments="Ornaments"
)

# "Transitory" first so it is the reference level: the interaction term then reads as
# permanent minus transitory, which is the direction the hypothesis is stated in.
marker_type_lookup <- c(
  circumcision          = "Permanent",
  tattoos_scarification = "Permanent",
  dress                 = "Transitory",
  food_taboos           = "Transitory",
  hair                  = "Transitory",
  ornaments             = "Transitory"
)

sample_dirs <- c(
  "Full sample"     = here("data","model","external","input"),
  "Markers present" = here("data","model","external_markers_present","input")
)

# stack the six per-marker files into one long frame: one row per (entry, marker)
read_long <- function(dir) {
  map_dfr(markers, function(m) {
    read_csv(file.path(dir, paste0(m, ".csv")), show_col_types = FALSE) |>
      rename(y = all_of(m)) |>
      mutate(marker = m)
  }) |>
    mutate(
      marker       = factor(marker, levels = markers),
      marker_type  = factor(marker_type_lookup[as.character(marker)],
                            levels = c("Transitory", "Permanent")),
      world_region = fct_relevel(factor(world_region), "Africa"),
      entry_id     = factor(entry_id)
    )
}

# lkj(2) mildly regularizes the marker intercept/slope correlation, which is estimated
# from only six markers; the rest match 1_external.Rmd and 6_conditional.Rmd
priors <- c(
  prior(normal(0, 1.5), class = Intercept),
  prior(normal(0, 1),   class = b),
  prior(exponential(1), class = sd),
  prior(lkj(2),         class = cor)
)

fits <- list()
for (s in names(sample_dirs)) {
  d <- read_long(sample_dirs[[s]])

  f <- bf(y ~ 1 + violent_external * marker_type + year_scaled +
            (1 | world_region) + (1 + violent_external | marker) + (1 | entry_id))

  fit <- brm(f, data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 4000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99, max_treedepth = 12),
             seed = 51252, backend = "cmdstanr", silent = 2)

  fits[[s]] <- list(fit = fit, data = d,
                    n_rows = nrow(d), n_entries = n_distinct(d$entry_id))
}

```

Convergence

```{r diagnostics}

diag_tbl <- map_dfr(names(fits), function(s) {
  fit <- fits[[s]]$fit
  tibble(
    Sample       = s,
    n_rows       = fits[[s]]$n_rows,
    n_entries    = fits[[s]]$n_entries,
    divergences  = sum(subset(nuts_params(fit), Parameter=="divergent__")$Value),
    max_rhat     = round(max(rhat(fit), na.rm=TRUE), 3),
    min_bulk_ess = round(min(summary(fit)$fixed[,"Bulk_ESS"]), 0),
    min_tail_ess = round(min(summary(fit)$fixed[,"Tail_ESS"]), 0)
  )
})
diag_tbl

```

The contrast. `b_violent_external` is the conflict slope for transitory markers (the
reference level); the interaction is permanent minus transitory; their sum is the slope
for permanent markers. Reported on the log-odds scale with a 95% credible interval, plus
the directional posterior probability that permanent exceeds transitory.

```{r contrast_logodds}

fmt <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)

qi <- function(x) fmt(median(x), quantile(x, 0.025), quantile(x, 0.975))

contrast_draws <- map_dfr(names(fits), function(s) {
  draws  <- as_draws_df(fits[[s]]$fit)
  transitory <- draws$b_violent_external
  difference <- draws$`b_violent_external:marker_typePermanent`

  tibble(Sample = s, .draw = seq_along(transitory),
         Transitory = transitory,
         Permanent  = transitory + difference,
         Difference = difference)
})

# intervals and probabilities are summarised separately and joined: doing both in one
# summarise() would overwrite the numeric Difference draws with their formatted string
# before the probabilities below could be computed from them
contrast_intervals <- contrast_draws |>
  group_by(Sample) |>
  summarise(across(c(Transitory, Permanent, Difference), qi), .groups = "drop")

contrast_probs <- contrast_draws |>
  group_by(Sample) |>
  summarise(
    `P(diff > 0)` = round(mean(Difference > 0), 2),
    # ROPE: probability the difference is smaller than a half unit of log odds in either
    # direction. Stated explicitly because "no difference" cannot be read off an interval
    # that merely overlaps zero.
    `P(|diff| < 0.5)` = round(mean(abs(Difference) < 0.5), 2),
    .groups = "drop"
  )

contrast_tbl <- left_join(contrast_intervals, contrast_probs, by = "Sample") |>
  mutate(Sample = fct_relevel(Sample, "Full sample", "Markers present")) |>
  arrange(Sample)

contrast_tbl

```

The same contrast on the probability scale. AME for each marker type is the difference in
predicted probability between conflict and no conflict, for a typical marker of that type
(re_formula = NA, so marker, entry and region effects are all excluded), at the mean year.

```{r contrast_ame}

ame_draws <- map_dfr(names(fits), function(s) {
  fit <- fits[[s]]$fit
  d   <- fits[[s]]$data

  # grouping columns must be present even though re_formula = NA drops their effects
  nd <- expand_grid(
    violent_external = c(0, 1),
    marker_type      = factor(c("Transitory", "Permanent"),
                              levels = c("Transitory", "Permanent")),
    year_scaled      = 0
  ) |>
    mutate(world_region = d$world_region[1], marker = d$marker[1], entry_id = d$entry_id[1])

  fit |> epred_draws(newdata = nd, re_formula = NA) |> ungroup() |>
    select(.draw, violent_external, marker_type, .epred) |>
    pivot_wider(names_from = violent_external, values_from = .epred, names_prefix = "cond") |>
    mutate(ame = cond1 - cond0) |>
    select(.draw, marker_type, ame) |>
    pivot_wider(names_from = marker_type, values_from = ame) |>
    mutate(Difference = Permanent - Transitory, Sample = s)
})

# split for the same reason as contrast_tbl above
ame_intervals <- ame_draws |>
  group_by(Sample) |>
  summarise(across(c(Transitory, Permanent, Difference), qi), .groups = "drop")

ame_probs <- ame_draws |>
  group_by(Sample) |>
  summarise(`P(diff > 0)` = round(mean(Difference > 0), 2), .groups = "drop")

ame_tbl <- left_join(ame_intervals, ame_probs, by = "Sample") |>
  mutate(Sample = fct_relevel(Sample, "Full sample", "Markers present")) |>
  arrange(Sample)

ame_tbl

```

Per-marker conflict slopes implied by the model, to check that the type contrast is not
being carried by a single marker. These are shrunk toward their type mean, so they will
be less spread out than the independent per-marker fits.

```{r marker_slopes}

slope_tbl <- map_dfr(names(fits), function(s) {
  fit <- fits[[s]]$fit
  draws <- as_draws_df(fit)

  map_dfr(markers, function(m) {
    slope <- draws$b_violent_external +
      draws[[paste0("r_marker[", m, ",violent_external]")]] +
      if (marker_type_lookup[m] == "Permanent") draws$`b_violent_external:marker_typePermanent` else 0

    tibble(Sample = s, Marker = marker_labels[m],
           Type = marker_type_lookup[m], slope = qi(slope))
  })
}) |>
  mutate(Sample = fct_relevel(Sample, "Full sample", "Markers present")) |>
  arrange(Sample, match(Marker, marker_labels[markers]))

slope_tbl

```

```{r contrast_plot}

p_contrast <- contrast_draws |>
  pivot_longer(c(Transitory, Permanent, Difference),
               names_to = "quantity", values_to = "value") |>
  mutate(quantity = fct_relevel(quantity, "Difference", "Permanent", "Transitory"),
         Sample   = fct_relevel(Sample, "Full sample", "Markers present")) |>
  ggplot(aes(x = value, y = quantity, fill = Sample)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               position = position_dodge(width = 0.6), scale = 0.9,
               color = "black", alpha = 0.7) +
  scale_y_discrete(expand = expansion(add = c(0.6, 1.0))) +
  scale_fill_manual(values = c("Full sample" = "#C8839E", "Markers present" = "#6B9AC4")) +
  labs(x = "Effect of external violent conflict (log odds)", y = NULL, fill = NULL) +
  theme_bw(base_size = 16) +
  theme(
    axis.title.x    = element_text(size = 17),
    axis.text.y     = element_text(size = 15),
    axis.text.x     = element_text(size = 14),
    legend.position = "top",
    plot.margin     = margin(12, 16, 12, 16)
  )

p_contrast

```
