---
title: "7_combined"
output: html_document
---

Pooled model over the six child markers of "Are extra-ritual in-group markers present"
(ERM): does external conflict favour permanent over transitory markers?

Only ERM = "Yes" entries are used, because child answers under ERM = "No" are inherited
zeros, not observations. This fits P(child | ERM = Yes, conflict); P(ERM | conflict) is
fitted in 1_external.Rmd.

Two models:

1. common slope within each marker type
2. A + marker-specific deviations in the conflict slope, (0 + violent_external | marker)

delta = violent_external:permanent (permanent minus transitory slope). Coefficients are
conditional on entry, so only delta is comparable to 1_external.Rmd / 6_conditional.Rmd.

```{r}
library(pacman)
p_load(brms, tidyverse, here)
set.seed(12532)

dir.create(here("data","model","external_markers_present","results"),
           recursive = TRUE, showWarnings = FALSE)
dir.create(here("data","model","external_markers_present","fits"),
           recursive = TRUE, showWarnings = FALSE)
```

Data

```{r data}

markers <- c(
  "circumcision", "tattoos_scarification",       # permanent
  "dress", "food_taboos", "hair", "ornaments"    # transitory
)
permanent_markers <- c("circumcision", "tattoos_scarification")

marker_labels <- c(
  circumcision="Circumcision", tattoos_scarification="Tattoos/Scarification",
  dress="Dress", food_taboos="Food Taboos", hair="Hair", ornaments="Ornaments"
)

# one row per entry x marker; inputs are already ERM = "Yes" and observed values only
d <- map_dfr(markers, function(m) {
  read_csv(here("data","model","external_markers_present","input", paste0(m, ".csv")),
           show_col_types = FALSE) |>
    rename(y = all_of(m)) |>
    mutate(marker = m)
})

d <- d |>
  filter(!is.na(y)) |>
  mutate(
    marker = factor(marker, levels = markers),
    permanent = if_else(marker %in% permanent_markers, 1, 0),
    world_region = fct_relevel(factor(world_region), "Africa"),
    entry_id = factor(entry_id)
  )

cat("entries:", n_distinct(d$entry_id), "\n")
cat("rows:   ", nrow(d), "\n")

count_tbl <- d |>
  group_by(marker) |>
  summarise(
    permanent = first(permanent),
    n = n(),
    n_yes = sum(y),
    n_conflict = sum(violent_external),
    .groups = "drop"
  ) |>
  mutate(marker = marker_labels[as.character(marker)])
count_tbl

```

Models

```{r fit}

priors <- c(
  prior(normal(0, 1.5), class = b, coef = markercircumcision),
  prior(normal(0, 1.5), class = b, coef = markertattoos_scarification),
  prior(normal(0, 1.5), class = b, coef = markerdress),
  prior(normal(0, 1.5), class = b, coef = markerfood_taboos),
  prior(normal(0, 1.5), class = b, coef = markerhair),
  prior(normal(0, 1.5), class = b, coef = markerornaments),
  prior(normal(0, 1),   class = b),
  prior(exponential(1), class = sd)
)

formulas <- list(
  A = bf(y ~ 0 + marker + violent_external + violent_external:permanent + year_scaled +
           (1 | entry_id) + (1 | world_region) + (1 | world_region:marker)),
  B = bf(y ~ 0 + marker + violent_external + violent_external:permanent + year_scaled +
           (1 | entry_id) + (1 | world_region) + (1 | world_region:marker) +
           (0 + violent_external | marker))
)

fits <- list()
for (model in names(formulas)) {
  fit <- brm(formulas[[model]], data = d, family = bernoulli("logit"), prior = priors,
             chains = 4, iter = 8000, warmup = 2000, cores = 4,
             control = list(adapt_delta = 0.99),
             seed = 51252, backend = "cmdstanr", silent = 2)

  saveRDS(fit, here("data","model","external_markers_present","fits",
                    paste0("combined_", model, ".rds")))
  fits[[model]] <- fit
}

```

Convergence (ESS includes the group-level SDs)

```{r diagnostics}

diag_tbl <- map_dfr(names(fits), function(model) {
  fit <- fits[[model]]
  fixed <- summary(fit)$fixed
  sds <- bind_rows(map(summary(fit)$random, as.data.frame))

  tibble(
    model = model,
    divergences = sum(subset(nuts_params(fit), Parameter == "divergent__")$Value),
    max_rhat = round(max(rhat(fit), na.rm = TRUE), 3),
    min_bulk_ess = round(min(c(fixed[, "Bulk_ESS"], sds[, "Bulk_ESS"])), 0),
    min_tail_ess = round(min(c(fixed[, "Tail_ESS"], sds[, "Tail_ESS"])), 0)
  )
})
diag_tbl
write_csv(diag_tbl, here("data","model","external_markers_present","results","combined_diagnostics.csv"))

summary(fits$A)
summary(fits$B)

```

Posterior predictive check: observed prevalence against its posterior predictive
interval, split by conflict status within each marker, one panel per model.

```{r ppc}

ppc <- map_dfr(names(fits), function(model) {
  yrep <- posterior_predict(fits[[model]], ndraws = 1000)
  conflict <- ifelse(d$violent_external == 1, "Yes", "No")

  map_dfr(markers, function(mk) {
    map_dfr(c("No", "Yes"), function(g) {
      idx <- which(as.character(d$marker) == mk & conflict == g)
      prop <- rowMeans(yrep[, idx, drop = FALSE])
      tibble(
        model = model,
        Marker = unname(marker_labels[mk]),
        conflict = g,
        observed = mean(d$y[idx]),
        lo = as.numeric(quantile(prop, 0.025)),
        hi = as.numeric(quantile(prop, 0.975))
      )
    })
  })
}) |>
  mutate(Marker = factor(Marker, levels = rev(unname(marker_labels[markers]))))

ggplot(ppc, aes(y = Marker, colour = conflict)) +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 3, alpha = 0.4,
                 position = position_dodge(width = 0.5)) +
  geom_point(aes(x = observed), size = 2,
             position = position_dodge(width = 0.5)) +
  facet_wrap(~ model) +
  scale_colour_manual(values = c(No = "#8FBF9F", Yes = "#D98B72")) +
  labs(x = "Marker prevalence", y = NULL, colour = "External conflict",
       subtitle = "Bars: 95% posterior predictive interval. Points: observed proportion.") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

```

Contrast. Posterior mean, 95% CI and PP > 0, constructed as in 1_external.Rmd.

```{r contrast}

contrasts <- c(
  "delta (permanent - transitory)" = "violent_external:permanent",
  "transitory slope" = "violent_external",
  "permanent slope" = "violent_external + violent_external:permanent"
)

contrast_tbl <- map_dfr(names(fits), function(model) {
  map_dfr(names(contrasts), function(label) {
    h <- hypothesis(fits[[model]], paste(contrasts[[label]], "> 0"), alpha = 0.025)$hypothesis

    tibble(
      model = model,
      quantity = label,
      mean = round(h$Estimate, 2),
      lower_95 = round(h$CI.Lower, 2),
      upper_95 = round(h$CI.Upper, 2),
      pp_gt_0 = round(h$Post.Prob, 3)
    )
  })
})

contrast_tbl
write_csv(contrast_tbl, here("data","model","external_markers_present","results","combined_contrast.csv"))

```
