---
title: "Sanity check: phylogenetic covariance matrix and signal"
output:
  html_document:
    toc: true
    toc_float: true
---

Three questions:

1. **Is A correct?** Do the pairwise correlations match our linguistic intuitions?
2. **Is there any phylogenetic signal in the data?** If the outcome and predictor are randomly scattered across the tree, the phylogenetic control won't change the coefficient — not because the model is broken, but because there is genuinely little phylogenetic clustering.
3. **What is the overall structure of A?** Are most pairwise correlations near zero (deep divergences dominate), or is there meaningful spread?

```{r setup, message=FALSE, warning=FALSE}
library(pacman)
p_load(brms, ape, tidyverse, here, HDInterval, phytools)
```

## 1. Load data and build A

```{r load, message=FALSE}
data <- read_csv(here("data", "phylo_input", "extra_ritual_group_markers.csv"))
tree <- read.tree(here("data", "jaeger2018", "world.tre"))

tips_in_data <- unique(data$tip_name)
tree_pruned  <- keep.tip(tree, tips_in_data)
A <- vcv.phylo(tree_pruned, corr = TRUE)

cat("Tips in pruned tree:", length(tree_pruned$tip.label), "\n")
cat("A dimensions:", dim(A), "\n")
cat("Diagonal (should all be 1):", all(round(diag(A), 10) == 1), "\n")
```

```{r}
library(ape)
length(tree$tip.label)   # how many languages? (n=6892)
head(tree$tip.label)     # what do the tip names look like?
summary(tree$edge.length)
```

```{r}
# If you've already built A:
A[1:4, 1:4]        # look at a small block
```



## 2. Spot checks — are expected pairs correlated correctly?

Languages within the same family should have high correlation; languages from
different major families should be near zero.

```{r spot_checks}

# Helpers
cor_pair <- function(a, b) {
    if (!a %in% rownames(A)) { cat(a, "NOT FOUND\n"); return(NA) }
    if (!b %in% rownames(A)) { cat(b, "NOT FOUND\n"); return(NA) }
    round(A[a, b], 3)
}
find_tip <- function(pattern) grep(pattern, rownames(A), value = TRUE)

# All-pairs correlation within a set of tips, returned as a tidy tibble
within_cors <- function(tips, label) {
    pairs <- combn(tips, 2, simplify = FALSE)
    tibble(
        group = label,
        tip1  = sapply(pairs, `[`, 1),
        tip2  = sapply(pairs, `[`, 2),
        cor   = sapply(pairs, \(p) cor_pair(p[1], p[2]))
    )
}

# ── families with enough tips for a thorough check ──────────────────────────
tips_romance <- find_tip("IE\\.ROMANCE")          # 7 tips
tips_indic   <- find_tip("IE\\.INDIC")            # 10 tips
tips_slavic  <- find_tip("IE\\.SLAVIC")           # 4 tips
tips_semitic <- find_tip("AA\\.SEMITIC")          # 13 tips
tips_egypt   <- find_tip("AA\\.EGYPTIAN")         # 2 tips
tips_nahuatl <- find_tip("UA\\.AZTECAN")          # 15 tips (all Nahuatl dialects)
tips_chinese <- find_tip("ST\\.CHINESE")          # 5 tips
tips_tibetan <- find_tip("ST\\.BODIC")            # 6 tips

# ── 1. within-subfamily (should be HIGH) ────────────────────────────────────
within_results <- bind_rows(
    within_cors(tips_romance, "Within IE.ROMANCE"),
    within_cors(tips_indic,   "Within IE.INDIC"),
    within_cors(tips_slavic,  "Within IE.SLAVIC"),
    within_cors(tips_semitic, "Within AA.SEMITIC"),
    within_cors(tips_nahuatl, "Within UA.AZTECAN (Nahuatl dialects)"),
    within_cors(tips_chinese, "Within ST.CHINESE"),
    within_cors(tips_tibetan, "Within ST.BODIC")
)

within_results |>
    group_by(group) |>
    summarise(n_pairs = n(), median_cor = round(median(cor), 3),
              min_cor = round(min(cor), 3), max_cor = round(max(cor), 3)) |>
    print(n = Inf)
```

```{r spot_checks2}
# ── 2. between-subfamily, same macro-family (should be INTERMEDIATE) ────────
cat("=== IE: Romance vs Indic (same IE, different branch) ===\n")
cross_ie <- cross_join(
    tibble(t1 = tips_romance),
    tibble(t2 = tips_indic)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_ie$cor), 3),
    "  Range:", round(min(cross_ie$cor), 3), "–", round(max(cross_ie$cor), 3), "\n")

cat("\n=== IE: Romance vs Slavic (same IE, different branch) ===\n")
cross_ie2 <- cross_join(
    tibble(t1 = tips_romance),
    tibble(t2 = tips_slavic)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_ie2$cor), 3),
    "  Range:", round(min(cross_ie2$cor), 3), "–", round(max(cross_ie2$cor), 3), "\n")

cat("\n=== AA: Semitic vs Egyptian-Coptic (same AA, different branch) ===\n")
cross_aa <- cross_join(
    tibble(t1 = tips_semitic),
    tibble(t2 = tips_egypt)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_aa$cor), 3),
    "  Range:", round(min(cross_aa$cor), 3), "–", round(max(cross_aa$cor), 3), "\n")

cat("\n=== ST: Chinese vs Bodic (same ST, different branch) ===\n")
cross_st <- cross_join(
    tibble(t1 = tips_chinese),
    tibble(t2 = tips_tibetan)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_st$cor), 3),
    "  Range:", round(min(cross_st$cor), 3), "–", round(max(cross_st$cor), 3), "\n")
```

```{r spot_checks3}
# ── 3. between macro-families (should be NEAR ZERO) ─────────────────────────
cat("=== IE (Romance) vs AA (Semitic) ===\n")
cross_ie_aa <- cross_join(
    tibble(t1 = tips_romance),
    tibble(t2 = tips_semitic)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_ie_aa$cor), 3),
    "  Range:", round(min(cross_ie_aa$cor), 3), "–", round(max(cross_ie_aa$cor), 3), "\n")

cat("\n=== IE (Romance) vs ST (Chinese) ===\n")
cross_ie_st <- cross_join(
    tibble(t1 = tips_romance),
    tibble(t2 = tips_chinese)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_ie_st$cor), 3),
    "  Range:", round(min(cross_ie_st$cor), 3), "–", round(max(cross_ie_st$cor), 3), "\n")

cat("\n=== IE (Romance) vs UA (Nahuatl) ===\n")
cross_ie_ua <- cross_join(
    tibble(t1 = tips_romance),
    tibble(t2 = tips_nahuatl)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_ie_ua$cor), 3),
    "  Range:", round(min(cross_ie_ua$cor), 3), "–", round(max(cross_ie_ua$cor), 3), "\n")

cat("\n=== AA (Semitic) vs UA (Nahuatl) ===\n")
cross_aa_ua <- cross_join(
    tibble(t1 = tips_semitic),
    tibble(t2 = tips_nahuatl)
) |> mutate(cor = map2_dbl(t1, t2, cor_pair))
cat("Median:", round(median(cross_aa_ua$cor), 3),
    "  Range:", round(min(cross_aa_ua$cor), 3), "–", round(max(cross_aa_ua$cor), 3), "\n")
```

## 3. Distribution of off-diagonal correlations

If the world tree is very deep (long branches separating major families), most
pairwise correlations will be near zero. This tells us how much leverage the
covariance structure actually has.

```{r cor_distribution}
# Extract upper triangle (off-diagonal)
upper <- A[upper.tri(A)]

tibble(correlation = upper) |>
    ggplot(aes(x = correlation)) +
    geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
    geom_vline(xintercept = median(upper), linetype = "dashed", colour = "red") +
    labs(title = "Distribution of pairwise phylogenetic correlations",
         subtitle = paste0("Median = ", round(median(upper), 3),
                           "   |   % above 0.1: ",
                           round(mean(upper > 0.1) * 100, 1), "%"),
         x = "Correlation (shared branch length / total depth)",
         y = "Count")

cat("Median off-diagonal correlation:", round(median(upper), 4), "\n")
cat("% of pairs with correlation > 0.1:", round(mean(upper > 0.1) * 100, 1), "\n")
cat("% of pairs with correlation > 0.5:", round(mean(upper > 0.5) * 100, 1), "\n")
```

## 4. Is there phylogenetic signal in the data?

Blomberg's K measures whether related languages are more similar in a trait than
expected by chance. K ≈ 1 = trait evolves by Brownian motion; K < 1 = less
clustering than expected; K > 1 = strong clustering. For binary traits this is
approximate, but useful directionally.

```{r phylo_signal}
# Aggregate traits to one value per tip (mean) for this diagnostic only
tip_means <- data |>
    group_by(tip_name) |>
    summarise(
        extra_ritual   = mean(extra_ritual_group_markers),
        violent_ext    = mean(violent_external),
        .groups = "drop"
    ) |>
    filter(tip_name %in% tree_pruned$tip.label)

# Named vectors matching tree tip order
ritual_vec  <- setNames(tip_means$extra_ritual,  tip_means$tip_name)
violent_vec <- setNames(tip_means$violent_ext,   tip_means$tip_name)

K_ritual  <- phylosig(tree_pruned, ritual_vec[tree_pruned$tip.label],  method = "K")
K_violent <- phylosig(tree_pruned, violent_vec[tree_pruned$tip.label], method = "K")

cat("Blomberg's K — extra_ritual_group_markers:", round(K_ritual,  3), "\n")
cat("Blomberg's K — violent_external:          ", round(K_violent, 3), "\n")
cat("\nInterpretation:\n")
cat("  K << 1: trait is scattered randomly across the tree -> phylo control has little effect\n")
cat("  K ~  1: Brownian motion signal\n")
cat("  K >> 1: strong family-level clustering\n")
```

## 5. Visualise the pruned tree

A quick plot to confirm the tree looks sensible and that tips from the same family
cluster together as expected.

```{r tree_plot, fig.height=14, fig.width=7}
# Colour tips by family (first component of tip label, e.g. "IE", "AA", "Sino")
family <- sub("\\..*", "", tree_pruned$tip.label)
n_fam  <- length(unique(family))

plot(tree_pruned, type = "fan", cex = 0.35, no.margin = TRUE,
     tip.color = as.integer(as.factor(family)),
     main = paste0("Pruned ASJP tree (", length(tree_pruned$tip.label), " tips)"))
```

## 6. What does this tell us about the stable coefficient?

```{r interpretation}
cat("Summary:\n\n")
cat("Median pairwise correlation in A:", round(median(upper), 3), "\n")
cat("Blomberg's K (outcome):  ", round(K_ritual,  3), "\n")
cat("Blomberg's K (predictor):", round(K_violent, 3), "\n\n")

cat("If K values are both low (< 0.3), the most likely explanation for the\n")
cat("stable coefficient is that both traits are phylogenetically random —\n")
cat("the association between violent conflict and ritual markers is not driven\n")
cat("by shared ancestry. That is actually a *stronger* result, not a failure\n")
cat("of the phylogenetic control.\n\n")

cat("If K is moderate but the coefficient is still stable, it means the\n")
cat("phylogenetic clustering in the outcome is not confounded with the predictor.\n")
```
